### 2023-06-24 manuscript/manuscript.qmd
initial commit and initial draft of manuscript introduction

---
title: "The FAIRification of pharmacoepidemiological research"
subtitle: "A practical introduction to reproducible analytical workflows using git and R"
# date: "`r Sys.Date()`"
# date-format: long
# author:
#   - name: Janick Weberpals, RPh, PhD
#     corresponding: true
#     orcid: 0000-0003-0404-7394
#     email: jweberpals@bwh.harvard.edu
#     url: janickweberpals.github.io
#     affiliations:
#       - name: Division of Pharmacoepidemiology and Pharmacoeconomics <br> Brigham and Women's Hospital <br> Harvard Medical School
#         address: 1620 Tremont Street
#         city: Boston
#         state: MA
#         postal-code: 02120
#   - name: Shirley V. Wang, PhD
#     corresponding: true
#     orcid: 0000-0003-0404-7394
#     email: svwang1@bwh.harvard.edu
#     affiliations:
#       - name: Division of Pharmacoepidemiology and Pharmacoeconomics <br> Brigham and Women's Hospital <br> Harvard Medical School
#         address: 1620 Tremont Street
#         city: Boston
#         state: MA
#         postal-code: 02120
code-fold: false
format: docx
---

Janick Weberpals^1^, Shirley V. Wang^1^

Author affiliations:

^1^ Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women’s Hospital, Harvard Medical School, Boston, MA

Correspondence:
Janick Weberpals, RPh, PhD
Instructor in Medicine
Division of Pharmacoepidemiology and Pharmacoeconomics,
Department of Medicine, Brigham and Women’s Hospital, Harvard Medical School,
1620 Tremont Street, Suite 3030-R, Boston, MA 02120, USA
Phone: 617-278-0932 
Fax: 617-232-8602 
Email: jweberpals@bwh.harvard.edu

Word count: ... words

Tables:

Figures:

Supplementary material:

### 2023-06-26 manuscript/manuscript.qmd
added bibliography referencing style and custom reference doc for compilation into word document

@@
 ---
-title: "The FAIRification of pharmacoepidemiological research"
-subtitle: "A practical introduction to reproducible analytical workflows using git and R"
-# date: "`r Sys.Date()`"
-# date-format: long
-# author:
-#   - name: Janick Weberpals, RPh, PhD
-#     corresponding: true
-#     orcid: 0000-0003-0404-7394
-#     email: jweberpals@bwh.harvard.edu
-#     url: janickweberpals.github.io
-#     affiliations:
-#       - name: Division of Pharmacoepidemiology and Pharmacoeconomics <br> Brigham and Women's Hospital <br> Harvard Medical School
-#         address: 1620 Tremont Street
-#         city: Boston
-#         state: MA
-#         postal-code: 02120
-#   - name: Shirley V. Wang, PhD
-#     corresponding: true
-#     orcid: 0000-0003-0404-7394
-#     email: svwang1@bwh.harvard.edu
-#     affiliations:
-#       - name: Division of Pharmacoepidemiology and Pharmacoeconomics <br> Brigham and Women's Hospital <br> Harvard Medical School
-#         address: 1620 Tremont Street
-#         city: Boston
-#         state: MA
-#         postal-code: 02120
+title: "The FAIRification of pharmacoepidemiological research: A practical introduction to reproducible analytical workflows using git and R"
+
 code-fold: false
-format: docx
+eval: false
+number-sections: true
+
+format: 
+  docx:
+    reference-doc: custom-reference-doc.docx
+editor: visual
+bibliography: references.bib
+csl: pharmacoepidemiology-and-drug-safety.csl
 ---
 
-Janick Weberpals^1^, Shirley V. Wang^1^
+**Authors**: Janick Weberpals^1^, Shirley V. Wang^1^
 
-Author affiliations:
+**Author affiliation**: ^1^Division of Pharmacoepidemiologyand Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
 
-^1^ Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women’s Hospital, Harvard Medical School, Boston, MA
+[**Correspondence:**]{.underline}
 
-Correspondence:
 Janick Weberpals, RPh, PhD
+
 Instructor in Medicine
+
 Division of Pharmacoepidemiology and Pharmacoeconomics,
-Department of Medicine, Brigham and Women’s Hospital, Harvard Medical School,
+
+Department of Medicine, Brigham and Women's Hospital, Harvard Medical School,
+
 1620 Tremont Street, Suite 3030-R, Boston, MA 02120, USA
+
 Phone: 617-278-0932
+
 Fax: 617-232-8602
-Email: jweberpals@bwh.harvard.edu
 
-Word count: ... words
+Email: [jweberpals\@bwh.harvard.edu](mailto:jweberpals@bwh.harvard.edu)
 
-Tables:
+[**Word count:**]{.underline} xxx words
 
-Figures:
+[**Tables:**]{.underline}
 
-Supplementary material:
+[**Figures:**]{.underline}
+
+[**Supplementary material:**]{.underline}
+
+[**Keywords:**]{.underline}
+
+[**Funding:**]{.underline}
+
+[**Disclosures/COI:**]{.underline} Janick Weberpals reports no conflicts of interest.
+
+[**Analytical code sharing statement:**]{.underline}Figures and exemplary analyses were conducted using R version 4.1.2.
+Detailed information on used packages and versions can be found in the Supplementary Methods and code used in this study is available at .xxx.
+
+[**Data sharing statement:**]{.underline} ...
+
+[**Acknowledgments:**]{.underline}
+
+{{< pagebreak >}}
+
+# Abstract {.unnumbered}
+
+**Objective**
+
+**Study Design and Settings**
+
+**Results**
+
+**Conclusions**
+
+{{< pagebreak >}}
+
+# Introduction
+
+Transparency and reproducibility in conducting healthcare database studies in pharmacoepidemiology is one of the most critical scientific requirements for meaningful research.
+While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], the transparency around the actual implementation and analytical steps is still significantly lacking, especially when analyses are implemented via line programming steps in statistical programming languages like R[@RCore2023] or Python.
+Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018] or health informatics[@bakken2019], the sharing of programming code is frequently observed, this practice is not as common in the (pharmaco-)epidemiological community.
+
+Here, the principles of *FAIR* *data*[@garcía-closas2023] can be extended to the implementation of pharmacoepidemiological study workflows by making all components of a study **F**indable, **A**ccessible, **I**nteroperable, and **R**eproducible (**FAIR**).
+Version control systems (VCS), such as *Git*, can thereby provide easy and powerful tools to keep track of the versioning of analytical code, tables, figures and other resources, track and resolve errors, collaborate with peers, and share resources instantaneously upon submission or publication of a manuscript.
+By design, VCS workflows automatically enable researchers to comment, track and compare major changes made to documents and hence, increase the ability to comprehend the evolution of a project while maintaining a complete audit trail of all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as \"*final_analysis_1_rev_more_changes.R*\" which is bad practice as this is highly error prone, untransparent for collaborating project members and outside personnel, and changes are difficult to track and reconcile.
+Particularly in disciplines like pharmacoepidemiology, where analyses can have far reaching impact on healthcare decisions, analytical code needs to be accessible and reproducible by anyone, not only the analyst who originally carried out the analysis.
+
+To increase the adoption of transparent and repdroducible workflows using VCS, this tutorial aims to give a practical and tailored introduction for pharmacoepidemiologists on how to set up, structure, and implement workflows using Git as the most frequently used VCS in the scientific community.[@blischak2016] We will first shortly introduce Git and how it can be used to for collaboration and dissemination of study results using project repositories.
+Then we will provide a technical step-by-step guidance on how to integrate Git in an analytical workflow using line commands.
+We will additionally showcase examples using the R open-source programming language, although the basic principles can be applied to any major coding language.
+
+{{< pagebreak >}}
+
+# Git basics
+
+Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects simply and efficiently.[@chacon2014pro]
+
+{{< pagebreak >}}
+
+# Discussion
+
+Advantages of VCS:
+
+-   Free
+
+-
+
+{{< pagebreak >}}
+
+# References {.unnumbered}

### 2023-06-26 manuscript/manuscript.qmd
initialized manuscript draft in quarto

@@
 eval: false
+echo: true
 number-sections: true
@@
 
-Here, the principles of *FAIR* *data*[@garcía-closas2023] can be extended to the implementation of pharmacoepidemiological study workflows by making all components of a study **F**indable, **A**ccessible, **I**nteroperable, and **R**eproducible (**FAIR**).
-Version control systems (VCS), such as *Git*, can thereby provide easy and powerful tools to keep track of the versioning of analytical code, tables, figures and other resources, track and resolve errors, collaborate with peers, and share resources instantaneously upon submission or publication of a manuscript.
-By design, VCS workflows automatically enable researchers to comment, track and compare major changes made to documents and hence, increase the ability to comprehend the evolution of a project while maintaining a complete audit trail of all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as \"*final_analysis_1_rev_more_changes.R*\" which is bad practice as this is highly error prone, untransparent for collaborating project members and outside personnel, and changes are difficult to track and reconcile.
+Version control systems (VCS), such as *Git*, can thereby provide easy and powerful tools to keep track of the versioning of important files and documents such as protocols, analytical code, tables and figures.
+Thereby, it can extend the principles of *FAIR* *data*[@garcía-closas2023] to the implementation of pharmacoepidemiological study workflows by making all components of a study **F**indable, **A**ccessible, **I**nteroperable, and **R**eproducible (**FAIR**).
+
+VCS enable researchers to track and resolve errors, collaborate with peers, and share resources instantaneously, for example, upon submission or publication of a manuscript.
+By design, VCS workflows automatically empower users to comment, track and compare changes made to files and hence, increase the ability to comprehend the evolution of a project over time while maintaining a complete audit trail of all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as \"*final_analysis_1_rev_more_changes.R*\" which is bad practice as this is highly error prone, lacking transparency for collaborating project members and outside personnel, and changes are difficult to track and reconcile.
 Particularly in disciplines like pharmacoepidemiology, where analyses can have far reaching impact on healthcare decisions, analytical code needs to be accessible and reproducible by anyone, not only the analyst who originally carried out the analysis.
 
-To increase the adoption of transparent and repdroducible workflows using VCS, this tutorial aims to give a practical and tailored introduction for pharmacoepidemiologists on how to set up, structure, and implement workflows using Git as the most frequently used VCS in the scientific community.[@blischak2016] We will first shortly introduce Git and how it can be used to for collaboration and dissemination of study results using project repositories.
+To increase the adoption of transparent and repdroducible workflows in real-world evidence (RWE) studies using VCS, this tutorial aims to give a practical and tailored introduction for pharmacoepidemiologists on how to set up, structure, and implement workflows using *Git*, which is the most frequently used VCS to date.[@blischak2016] We will first provide a brief introduction to Git and how it can be used to for collaboration and dissemination of study results using project repositories.
 Then we will provide a technical step-by-step guidance on how to integrate Git in an analytical workflow using line commands.
@@
 
-# Git basics
+# Git in a nutshell
 
-Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects simply and efficiently.[@chacon2014pro]
+Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it it makes a *snapshot* of a research project repository every time the state of a project is saved.
+For our purposes, a repository can be seen as an isolated project directory which includes all neccessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).
+
+In its basic form, Git is a local software and does not need any internet connectivity or connection to a remote server to store *local* changes made to files in a repository.
+However, to leverage the full potential of Git, it is almost always used in combination with web-based remote repository hosting services such as *GitHub*, *GitLab*, *Bitbucket* and other platforms (henceforth referred to as remote repositories, **Overview Box 1**).
+
+::: {.callout-tip}
+## **Overview Box 1**. Differences between Git and remote repository platforms (GitHub, GitLab, etc.)
+
+**Git**
+
+-   Is an open-source software and distributed version control system
+
+-   Generally operated through command line tools, graphical user interfaces are available
+
+-   Installed locally on a system to track local changes of a Git repository
+
+**Remote repository platforms (GitHub, GitLab, etc.)**
+
+-   Web-based services, usually provided by a commercial entity
+
+-   Usually free for basic functions, paid services for more advanced features
+
+-   Needed to collaborate on shared repositories and dissemination of code
+
+-   Offer a graphical user interface and other functionalities that complement basic Git functions
+
+:::
+
+These platforms make it possible to synchronize local Git repositories of multiple project members and thereby enable collaboration.
+They further complement Git functionalities by providing a graphical user interface to visualize changes made to files and often offer advanced features such as the automation of workflows using continuous integration and deployment (CI/CD) or, most recently, AI-powered virtual coding assistants (e.g., GitHub co-pilot).
+Generally speaking, it's possible to use Git without a remote repository, while it's unusual to use a remote repository without Git.
+
+# A step-by-step introduction on how to use Git in a pharmacoepidemiology research study
+
+In this section, we want to give a technical introduction on how Git and remote repositories can be used in RWE studies.
+For the following examples, we focus on workflows utilizing R/RStudio ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitLab (<https://about.gitlab.com/>) as the remote repository, although the general concepts are similar and easily transferable to other programming languages and remote repository providers, respectively.
+We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated graphical user interface (GUI).
+For a most comprehensive collection of details, we refer to the book Pro Git by Chacon and Straub.[@chacon2014pro]
+
+## Installing Git
+
+Git is available for every contemporary operating system and can be downloaded on the official Git website: <https://git-scm.com/book/en/v2/Getting-Started-Installing-Git>.
+
+To check if the installation succeeded and show the available version:
+
+```{bash}
+git --version
+```
+
+## Configuration and first time use
+
+For Git to be able to track who made changes, the first step is to provide a name and email address as shown below. This needs to be done only once if the `--global` parameter is set.
+
+```{bash}
+git config --global user.name "First Last"
+git config --global user.email "name@domain.edu"
+```
+
+## Initialization of a new Git-controlled repository
+
+### Local initialization
+
+To start a new project (e.g., research study), the next step is to create and navigate to a new directory where all documents will be stored that are intended to be tracked via Git (in the following referred to as repository).
+In this example, the study repository is called `rwe_study`.
+
+```{bash}
+mkdir ~/rwe_study
+cd ~/rwe_study
+```
+
+The command to initialize this directory as a Git repository is
+
+```{bash}
+git init
+```
+
+This triggers the creation of the `.git` folder which will contain the entire version history of the `rwe_study` repository.
+
+### GUI initialization and `git clone`
+
+An alternative to creating a new repository is through creating a new project on the remote repository (in this case GitLab) and creating a local copy, a process which is usually referred to as `cloning`.
+
+To do so, you need to log into you remote repository profile and navigate to the section `New project/repository` \> `Create blank project` (or use a template) \> Enter some basic information of your new project such as the project name (`rwe_study`) and determine if the repository should be private or public and if a default `README` file should be created.
+
+Now it is possible to copy this pre-populated repository by copying the link given under `Clone` \> `HTTPS` into the CLI:
+
+```{bash}
+mkdir ~/rwe_study
+cd ~/rwe_study
+# GitHub example
+git clone https://github.com/USERNAME/rwe_study.git
+# GitLab example
+git clone  https://gitlab.com/USERNAME/rwe_study.gitlab.io.git
+```
+
+This can also be achieved directly in the RStudio GUI via `File` \> `New Project` \> `Version Control` \> `Git` \> Paste the link to `Repository URL` (fig...)
+
+commit == snapshot
+
+## Repository structure
+
+At this step, we can start now filling the repository with study-relevant documents. To enhance clarity, it is recommended to use
+
+# Discussion
+
+Advantages of distributed VCS (Git):
+
+-   Full audit trail of which changes were made by whom, when and why
+
+-   Even if the code is not shared, Git makes it easier for an individual to go back to the code
+
+-   Git can serve as internal and external repository for code and made accessible to everyone, e.g. along with manuscript without copy-paste into word files or similar
+
+-   It is free and open-source and reliable
+
+-   Distributed VCS fully mirror entire repositories and hence every local clone is a full backup of the data
 
@@
 
-# Discussion
+# Figures
 
-Advantages of VCS:
-
--   Free
-
--
+![Figure 2. Minimal example of a transparent repository structure containing relevant study documents.](figures/Figure_2_repo_structure.png "#fig-2-structure"){#fig-2-structure}
 

### 2023-06-27 manuscript/supplement.qmd
started a supplement file

---
title: "Supplementary Appendix"
subtitle: "The FAIRification of pharmacoepidemiological research: A practical introduction to reproducible analytical workflows using git and R"
code-fold: true
eval: false
echo: true
number-sections: true

format: 
  pdf:
    fig-cap-location: top
    crossref:
    fig-title: "**Supplementary Figure**"

editor: visual
bibliography: references.bib
csl: pharmacoepidemiology-and-drug-safety.csl
---

```{r setup}
#| label: setup
#| include: false
#| eval: true
library(magick)
library(cowplot)
library(here)
```

**Authors**: Janick Weberpals^1^, Shirley V. Wang^1^

**Author affiliation**: ^1^Division of Pharmacoepidemiologyand Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA

# Supplementary Figures

```{r}
#| label: fig-GitHub-gui_1
#| fig-cap: "Starting a new repository through the GitHub GUI."
#| eval: false
#| echo: false
#| out-width: 60%

sfig_a <- ggdraw() + 
  draw_image(image_read(here("manuscript/figures", "Figure_1a_new_repo.png")), scale = 0.5)

sfig_a
```

{{< pagebreak >}}

```{r}
#| label: fig-GitHub-gui
#| fig-cap: ""
#| eval: false
#| echo: false
#| out-width: 60%

sfig_a <- ggdraw() + 
  draw_image(image_read(here("manuscript/figures", "Figure_1a_new_repo.png")), scale = 0.5)

sfig_b <- ggdraw() + 
  draw_image(image_read(here("manuscript/figures", "Figure_1b_repo_info.png")))

sfig_c <- ggdraw() + 
  draw_image(image_read(here("manuscript/figures", "Figure_1c_clone.png")))

sfig1 <- plot_grid(
  fig_a, fig_c, fig_b, 
  nrow = 2,
  ncol = 2
)

sfig1
```

{{< pagebreak >}}

# References {.unnumbered}

### 2023-06-27 manuscript/manuscript.qmd
added figures and figure cross references and worked a bit on text

@@
     reference-doc: custom-reference-doc.docx
+    fig-cap-location: top
+
 editor: visual
@@
 ---
+
+```{r setup}
+#| label: setup
+#| include: false
+#| eval: true
+library(magick)
+library(cowplot)
+library(here)
+```
 
@@
 Janick Weberpals, RPh, PhD
-
-Instructor in Medicine
 
@@
 
-Transparency and reproducibility in conducting healthcare database studies in pharmacoepidemiology is one of the most critical scientific requirements for meaningful research.
+Transparency and reproducibility in conducting healthcare database studies in pharmacoepidemiology are critical scientific requirements for meaningful research.
 While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], the transparency around the actual implementation and analytical steps is still significantly lacking, especially when analyses are implemented via line programming steps in statistical programming languages like R[@RCore2023] or Python.
-Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018] or health informatics[@bakken2019], the sharing of programming code is frequently observed, this practice is not as common in the (pharmaco-)epidemiological community.
+Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018] or health informatics[@bakken2019], the sharing of programming code is rather the rule than the exception, this practice is not as common in the (pharmaco-)epidemiological community.
 
@@
 VCS enable researchers to track and resolve errors, collaborate with peers, and share resources instantaneously, for example, upon submission or publication of a manuscript.
-By design, VCS workflows automatically empower users to comment, track and compare changes made to files and hence, increase the ability to comprehend the evolution of a project over time while maintaining a complete audit trail of all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as \"*final_analysis_1_rev_more_changes.R*\" which is bad practice as this is highly error prone, lacking transparency for collaborating project members and outside personnel, and changes are difficult to track and reconcile.
+By design, VCS workflows automatically empower users to comment, track and compare changes made to files and hence, increase the ability to comprehend the evolution of a project over time while maintaining a complete audit trail of all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as "*final_analysis_1_rev_more_changes.R*" which is bad practice as this is highly error prone, lacking transparency for collaborating project members and outside personnel, and changes are difficult to track and reconcile.
 Particularly in disciplines like pharmacoepidemiology, where analyses can have far reaching impact on healthcare decisions, analytical code needs to be accessible and reproducible by anyone, not only the analyst who originally carried out the analysis.
 
-To increase the adoption of transparent and repdroducible workflows in real-world evidence (RWE) studies using VCS, this tutorial aims to give a practical and tailored introduction for pharmacoepidemiologists on how to set up, structure, and implement workflows using *Git*, which is the most frequently used VCS to date.[@blischak2016] We will first provide a brief introduction to Git and how it can be used to for collaboration and dissemination of study results using project repositories.
+To increase the adoption of transparent and repdroducible workflows in real-world evidence (RWE) studies using VCS, this tutorial aims to give a practical introduction tailored for pharmacoepidemiologists on how to set up, structure, and implement workflows using *Git*, which is the most frequently used VCS to date.[@blischak2016] We will first provide a brief introduction to Git and how it can be used to for collaboration and dissemination of study results using project repositories.
 Then we will provide a technical step-by-step guidance on how to integrate Git in an analytical workflow using line commands.
@@
 
-::: {.callout-tip}
+::: callout-tip
 ## **Overview Box 1**. Differences between Git and remote repository platforms (GitHub, GitLab, etc.)
@@
 -   Offer a graphical user interface and other functionalities that complement basic Git functions
-
 :::
 
-These platforms make it possible to synchronize local Git repositories of multiple project members and thereby enable collaboration.
-They further complement Git functionalities by providing a graphical user interface to visualize changes made to files and often offer advanced features such as the automation of workflows using continuous integration and deployment (CI/CD) or, most recently, AI-powered virtual coding assistants (e.g., GitHub co-pilot).
+These platforms make it possible to synchronize local Git repositories across multiple project members and thereby enable collaboration.
+They further complement Git functionalities by providing a graphical user interface (GUI) to visualize changes made to files and often offer advanced features such as the automation of workflows using continuous integration and deployment (CI/CD) or, most recently, AI-powered virtual coding assistants (e.g., GitHub co-pilot).
 Generally speaking, it's possible to use Git without a remote repository, while it's unusual to use a remote repository without Git.
@@
 In this section, we want to give a technical introduction on how Git and remote repositories can be used in RWE studies.
-For the following examples, we focus on workflows utilizing R/RStudio ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitLab (<https://about.gitlab.com/>) as the remote repository, although the general concepts are similar and easily transferable to other programming languages and remote repository providers, respectively.
-We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated graphical user interface (GUI).
-For a most comprehensive collection of details, we refer to the book Pro Git by Chacon and Straub.[@chacon2014pro]
+For the following examples, we focus on workflows utilizing R/RStudio ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as the remote repository, although the general concepts are similar and easily transferable to other programming languages and remote repository providers, respectively.
+We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated GUI.
+For more details, we refer to the book *Pro Git* by Chacon and Straub.[@chacon2014pro]
 
@@
 
-To check if the installation succeeded and show the available version:
+To check if the installation succeeded and show the available version, the following prompt in the CLI can be used.
 
@@
 
-To start a new project (e.g., research study), the next step is to create and navigate to a new directory where all documents will be stored that are intended to be tracked via Git (in the following referred to as repository).
+To start a new project (e.g., a RWE study), the next step is to create and navigate to a new directory where all documents that are intended to be tracked via Git will be stored (in the following referred to as repository).
 In this example, the study repository is called `rwe_study`.
@@
 
-This triggers the creation of the `.git` folder which will contain the entire version history of the `rwe_study` repository.
+This initiates the creation of the `.git` sub-directory within the study repository which will contain the entire version history of the `rwe_study` project.
+Typically, users don't directly interact with this sub-directory and the only thing to keep in mind is that this folder keeps all files needed to track changes.
 
-### GUI initialization and `git clone`
+### GUI initialization and `git clone` {#sec-gui-initialization-and-git-clone}
 
-An alternative to creating a new repository is through creating a new project on the remote repository (in this case GitLab) and creating a local copy, a process which is usually referred to as `cloning`.
+An alternative to creating a new repository locally is through creating a new project in the remote repository (in this case GitHub) and then creating a local copy, a process which is usually referred to as `cloning`.
 
-To do so, you need to log into you remote repository profile and navigate to the section `New project/repository` \> `Create blank project` (or use a template) \> Enter some basic information of your new project such as the project name (`rwe_study`) and determine if the repository should be private or public and if a default `README` file should be created.
+To do so, users will need to log into their remote repository account and navigate to the section `New repository` (**Supplementary Figure 1**).
+This will take a user to a form to populate basic information of the new project such as the repository name (`rwe_study`) and choose basic setting such as if the repository should be private or public and if a `README.md` file (see @sec-repository-structure) should be automatically created (**Supplementary Figure 2**).
 
-Now it is possible to copy this pre-populated repository by copying the link given under `Clone` \> `HTTPS` into the CLI:
+Now we can copy the pre-populated repository to our local machine by copying the link given under `Clone` \> `Copy HTTPS` into the CLI (**Supplementary Figure 3**):
 
@@
 cd ~/rwe_study
-# GitHub example
-git clone https://github.com/USERNAME/rwe_study.git
-# GitLab example
 git clone  https://gitlab.com/USERNAME/rwe_study.gitlab.io.git
@@
 
-This can also be achieved directly in the RStudio GUI via `File` \> `New Project` \> `Version Control` \> `Git` \> Paste the link to `Repository URL` (fig...)
+This can also be achieved directly in the RStudio GUI via `File` \> `New Project` \> `Version Control` \> `Git` \> paste the link to `Repository URL` (@fig-RStudio-gui).
 
-commit == snapshot
+## Repository structure {#sec-repository-structure}
 
-## Repository structure
+At this step, we now can populate the repository and start working on study-relevant documents.
+To enhance clarity, it is recommended to use a logical structure and informative and transparent sub-directory and file names (a minimal example is illustrated in @fig-structure).
+To document the structure and location of files, we can include this and other critical information that a peer or outside personnel should know concerning your project in the `README.md` file.
+This file can be auto-generated when a project is initialized in a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be added manually by creating and adding a text file with the name `README.md` in the root directory of the repository.
+We provide a pre-populated template of such a study repository based on the HARPER protocol [@wang2022] under ...
+.
 
-At this step, we can start now filling the repository with study-relevant documents. To enhance clarity, it is recommended to use
+## The git workflow
+
+The most essential steps of a git workflow are summarized in @fig-workflow.
+
+### Stage and commit changes
+
+As introduced before, Git allows researchers to keep track of changes by making snapshots of the repository every time the state of a project is saved.
+In Git terms, this fundamental step is referred to as `commit`.
+Generally it is up to the user which files should be committed, how the commit should be documented in a `commit message` and when it should be performed.
+
+To determine which files should be selected (or `staged`) for a new commit, the `git add` command is used followed by the name(s) of the files that should be committed or a `.` to stage all changes present in the current repository.
+
+```{bash}
+# Examplary staging command
+git add 'scripts/03_propensity_score_analysis.R'
+```
+
+Once all files that should be committed in the same step are staged, the `git commit` command will create a local snapshot of the changes.
+The `-m` suffix and quoted text after the command represent the commit message, i.e., an informative yet brief comment on what changes were performed.
+
+```{bash}
+# Examplary commit command
+git commit -m 'Applied inclusion and exclusion criteria to cohort query'
+```
+
+If any file(s) in the study repository should not be tracked and hence not be synchronized with the remote repository (e.g., because they include confidential information that should not be visible to others like database credentials), these files can be specified in the `.gitignore` file which needs to be located in the root directory of the repository.
+To create a `.gitignore` file, either auto-generate it upon initialization of the repository on GitHub (**Supplementary Figure 2**) or manually add a text file called `.gitignore` to the root directory.
+
+@fig-stage-commit-RStudio-gui illustrates the equivalent steps in using the RStudio GUI.
+
+## Synchronize changes with remote repository
+
+### Local changes to remote (`push`)
+
+To synchronize the local changes to our files, the next step involves synchronizing our commits with the remote repository on GitHub via the `push` command.
+
+```{bash}
+git push
+```
+
+In case we initialized the local repository as described in section @sec-gui-initialization-and-git-clone, a remote was already automatically configured. This can be checked via the following command.
+
+```{bash}
+git remote -v
+```
+
+If there is no remote established yet, this can be configured using the following commands.
+
+```{bash}
+git remote add origin https://github.com/USERNAME/rwe_study.git
+git branch -M main
+git push -u origin main
+```
+
+This step is not needed if the intention of a researcher is just to keep track of local changes and there is no intention to share code, collaborate with others or work across different systems.
+However, most of the times researchers need to collaborate on a project and `git` can improve collaborative work significantly (@fig-collaborate).
+
+### Remote to local
+
+In case collaborators have also made modification to their local files and pushed them to the remote repository, these changes can be easily synchronized with our local copy via the `pull` command.
+
+```{bash}
+git pull
+```
+
+Occasionally it can occur that two collaborators made conflicting changes to the same file or even to the same line of code.
+While Git is generally good at automatically integrating (`merging`) new changes, Git will notify the user to resolve a conflict by editing the conflicting file if Git cannot automatically determine what is correct.
+
+## Tracking changes
+
+## Keeping track of R software versions and paths
 
 # Discussion
+
+Also other GUIs to interact with Git...
+
+Docker
 
@@
 
+# Tables
+
+Create a table with most important git commands
+
 # Figures
 
-![Figure 2. Minimal example of a transparent repository structure containing relevant study documents.](figures/Figure_2_repo_structure.png "#fig-2-structure"){#fig-2-structure}
+```{r}
+#| label: Fig1
+#| fig-cap: "Clone a remote repository using the RStudio graphical user interface."
+#| eval: false
+#| echo: false
+
+fig1a <- ggdraw() +
+  draw_image(image_read(here("figures", "Figure_1a_RStudio.png")))
+
+fig1b <- ggdraw() +
+  draw_image(image_read(here("figures", "Figure_1b_RStudio.png")))
+
+fig1c <- ggdraw() +
+  draw_image(image_read(here("figures", "Figure_1c_RStudio.png")))
+
+fig1d <- ggdraw() +
+  draw_image(image_read(here("figures", "Figure_1d_RStudio.png")))
+
+figure_1 <- plot_grid(
+  fig1a, fig1b, fig1c, fig1d,
+  nrow = 2,
+  ncol = 2,
+  #labels = c("a)", "b)", "c)", "d)"),
+  hjust = 0.005,
+  label_size = 16
+  )
+
+png(
+  filename = here("figures", "Figure_1_RStudio_init.png"), 
+  width = 15, height = 10, units = "in", res=600
+  )
+figure_1
+invisible(dev.off())
+
+knitr::include_graphics(here::here("figures", "Figure_1_RStudio_init.png"))
+```
+
+```{r}
+#| label: fig-RStudio-gui
+#| fig-cap: "Clone a remote repository using the RStudio graphical user interface."
+#| eval: true
+#| echo: false
+#| out-width: 50%
+knitr::include_graphics(here::here("figures", "Figure_1_RStudio_init.png"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-structure
+#| fig-cap: "Minimal example of a transparent repository structure containing relevant study documents"
+#| eval: true
+#| echo: false
+#| out-width: 60%
+knitr::include_graphics(here::here("figures", "Figure_2_repo_structure.png"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-workflow
+#| fig-cap: "Overview of a basic Git workflow."
+#| eval: true
+#| echo: false
+knitr::include_graphics(here::here("figures", "Figure_3_workflow.png"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-stage-commit-RStudio-gui
+#| fig-cap: "RStudio provides a graphical user interface as an alternative to command line prompts to interact with Git and to perform command equivalent to git add, git commit and git push/pull (red circles).
+This view can be accessed in the right upper pane of RStudio under Git > Diff.
+In this example, an R script with the title '03_propensity_score_analysis.R' is selected and staged.
+All changes performed in this file can be viewed in the window below with green highlighted lines indicating additions and red highlighted lines deletions from the prior version.
+In this case, a title was added to a propensity score plot which is also expressed in the commit message."
+#| eval: true
+#| echo: false
+knitr::include_graphics(here::here("figures", "Figure_4_stage_commit.png"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-collaborate
+#| fig-cap: "Git in connection with remote repositories (e.g., GitHub or GitLab) can signficantly improve collaboration across project members."
+#| eval: true
+#| echo: false
+knitr::include_graphics(here::here("figures", "Figure_5_collaboration.png"))
+```
 

### 2023-06-27 manuscript/supplement.qmd
started a supplementary document

@@
 code-fold: true
-eval: false
-echo: true
+echo: false
 number-sections: true
@@
     fig-cap-location: top
+    fig-align: center
     crossref:
       fig-title: "**Supplementary Figure**"
+    toc: true
+    toc-depth: 2
+    number-sections: false
+    colorlinks: true
 
@@
 library(here)
+library(knitr)
 ```
 
-**Authors**: Janick Weberpals^1^, Shirley V. Wang^1^
-
-**Author affiliation**: ^1^Division of Pharmacoepidemiologyand Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
+{{< pagebreak >}}
 
@@
 ```{r}
-#| label: fig-GitHub-gui_1
+#| label: fig-GitHub-gui_a
 #| fig-cap: "Starting a new repository through the GitHub GUI."
-#| eval: false
-#| echo: false
-#| out-width: 60%
+#| out-width: 115%
 
-sfig_a <- ggdraw() + 
-  draw_image(image_read(here("manuscript/figures", "Figure_1a_new_repo.png")), scale = 0.5)
-
-sfig_a
+include_graphics(here("manuscript/figures", "SFigure_1a_new_repo.png"))
 ```
 
-{{< pagebreak >}}
+```{r}
+#| label: fig-GitHub-gui_b
+#| fig-cap: "Populating basic information about the new repository."
+#| out-width: 110%
+
+include_graphics(here("manuscript/figures", "SFigure_2_repo_info.png"))
+```
 
 ```{r}
-#| label: fig-GitHub-gui
-#| fig-cap: ""
-#| eval: false
-#| echo: false
-#| out-width: 60%
+#| label: fig-GitHub-gui_c
+#| fig-cap: "Clone a GitHub repository using the HTTPS link and the git clone command"
+#| out-width: 115%
 
-sfig_a <- ggdraw() + 
-  draw_image(image_read(here("manuscript/figures", "Figure_1a_new_repo.png")), scale = 0.5)
-
-sfig_b <- ggdraw() + 
-  draw_image(image_read(here("manuscript/figures", "Figure_1b_repo_info.png")))
-
-sfig_c <- ggdraw() + 
-  draw_image(image_read(here("manuscript/figures", "Figure_1c_clone.png")))
-
-sfig1 <- plot_grid(
-  fig_a, fig_c, fig_b, 
-  nrow = 2,
-  ncol = 2
-)
-
-sfig1
+include_graphics(here("manuscript/figures", "SFigure_3_clone.png"))
 ```

### 2023-06-28 manuscript/manuscript.qmd
added link to harper template repository
intermediate commit after finishing introduction and main part of the manuscript. Next up: discussion

@@
 ---
-title: "The FAIRification of pharmacoepidemiological research: A practical introduction to reproducible analytical workflows using git and R"
+title: "The FAIRification of pharmacoepidemiological research: A practical introduction to reproducible analytic workflows using git and R"
 
@@
 
-**Author affiliation**: ^1^Division of Pharmacoepidemiologyand Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
+**Author affiliation**: ^1^Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA
 
@@
 
-[**Word count:**]{.underline} xxx words
+[**Word count:**]{.underline} xxxx words
 
-[**Tables:**]{.underline}
+[**Tables:**]{.underline} 0
 
-[**Figures:**]{.underline}
+[**Figures:**]{.underline} 6
 
@@
 
-[**Keywords:**]{.underline}
+[**Keywords:**]{.underline} Transparency, Reproducibility, Git, R, Version control, Real-world evidence
 
-[**Funding:**]{.underline}
+[**Funding:**]{.underline} There was no specific funding for this manuscript.
 
@@
 
-[**Analytical code sharing statement:**]{.underline}Figures and exemplary analyses were conducted using R version 4.1.2.
-Detailed information on used packages and versions can be found in the Supplementary Methods and code used in this study is available at .xxx.
+[Analytical code sharing statement:]{.underline} This manuscript was written using `Quarto` version 1.3.353 (<https://quarto.org/>) and R version 4.1.2.
+All materials can be found at <https://gitlab-scm.partners.org/drugepi/fair-epi>.
+Detailed information on packages and versions can be found in the `renv.lock` file in the linked repository.
 
-[**Data sharing statement:**]{.underline} ...
-
-[**Acknowledgments:**]{.underline}
+[**Acknowledgments:**]{.underline} The authors would like to thank ...
 
@@
 
-To increase the adoption of transparent and repdroducible workflows in real-world evidence (RWE) studies using VCS, this tutorial aims to give a practical introduction tailored for pharmacoepidemiologists on how to set up, structure, and implement workflows using *Git*, which is the most frequently used VCS to date.[@blischak2016] We will first provide a brief introduction to Git and how it can be used to for collaboration and dissemination of study results using project repositories.
-Then we will provide a technical step-by-step guidance on how to integrate Git in an analytical workflow using line commands.
-We will additionally showcase examples using the R open-source programming language, although the basic principles can be applied to any major coding language.
+To increase the adoption of transparent and reproducible workflows in real-world evidence (RWE) studies using VCS, this tutorial aims to give a practical introduction tailored for pharmacoepidemiologists on how to set up, structure, and implement workflows using *Git*, which is the most frequently used VCS to date.[@blischak2016] We will first give a brief introduction to Git and how it can be used for collaboration and dissemination of study results using project repositories.
+Then we will provide a technical step-by-step guidance on how to integrate Git in an analytic workflow.
+We will additionally discuss and showcase important aspects of reproducibillity using the R open-source programming language, although the basic principles are applicable to any major coding language.
 
@@
 
-Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it it makes a *snapshot* of a research project repository every time the state of a project is saved.
+Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it makes a *snapshot* of a research project repository every time the state of a project is saved.
 For our purposes, a repository can be seen as an isolated project directory which includes all neccessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).
@@
 In its basic form, Git is a local software and does not need any internet connectivity or connection to a remote server to store *local* changes made to files in a repository.
-However, to leverage the full potential of Git, it is almost always used in combination with web-based remote repository hosting services such as *GitHub*, *GitLab*, *Bitbucket* and other platforms (henceforth referred to as remote repositories, **Overview Box 1**).
+However, to leverage the full potential of Git, it is almost always used in combination with web-based remote repository hosting services such as *GitHub*, *GitLab*, *Bitbucket* or other platforms (henceforth referred to as remote repositories, **Overview Box 1**).
 
@@
 
-These platforms make it possible to synchronize local Git repositories across multiple project members and thereby enable collaboration.
+These platforms make it possible to synchronize local Git repositories across multiple project members and thereby improve collaboration.
 They further complement Git functionalities by providing a graphical user interface (GUI) to visualize changes made to files and often offer advanced features such as the automation of workflows using continuous integration and deployment (CI/CD) or, most recently, AI-powered virtual coding assistants (e.g., GitHub co-pilot).
@@
 In this section, we want to give a technical introduction on how Git and remote repositories can be used in RWE studies.
-For the following examples, we focus on workflows utilizing R/RStudio ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as the remote repository, although the general concepts are similar and easily transferable to other programming languages and remote repository providers, respectively.
+For the following examples, we focus on workflows utilizing R/RStudio ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository providers, respectively.
 We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated GUI.
@@
 
-To start a new project (e.g., a RWE study), the next step is to create and navigate to a new directory where all documents that are intended to be tracked via Git will be stored (in the following referred to as repository).
+To start a new project (e.g., a new RWE study), the next step is to create and navigate to a new directory where all documents that are intended to be tracked via Git will be stored (in the following referred to as *repository*).
 In this example, the study repository is called `rwe_study`.
@@
 To do so, users will need to log into their remote repository account and navigate to the section `New repository` (**Supplementary Figure 1**).
-This will take a user to a form to populate basic information of the new project such as the repository name (`rwe_study`) and choose basic setting such as if the repository should be private or public and if a `README.md` file (see @sec-repository-structure) should be automatically created (**Supplementary Figure 2**).
+This will take a user to a form to populate basic information of the new project such as the repository name (`rwe_study`) and choose basic setting such as if the repository should be private or public and if a `README.md` (see @sec-repository-structure) and `.gitignore` (see @sec-stage-and-commit-changes) file should be automatically created (**Supplementary Figure 2**).
 
-Now we can copy the pre-populated repository to our local machine by copying the link given under `Clone` \> `Copy HTTPS` into the CLI (**Supplementary Figure 3**):
+Now we can copy the pre-populated repository to our local machine by copy-pasting the link given under `Clone` \> `Copy HTTPS` into the CLI (**Supplementary Figure 3**):
 
@@
 To enhance clarity, it is recommended to use a logical structure and informative and transparent sub-directory and file names (a minimal example is illustrated in @fig-structure).
-To document the structure and location of files, we can include this and other critical information that a peer or outside personnel should know concerning your project in the `README.md` file.
-This file can be auto-generated when a project is initialized in a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be added manually by creating and adding a text file with the name `README.md` in the root directory of the repository.
-We provide a pre-populated template of such a study repository based on the HARPER protocol [@wang2022] under ...
-.
+To document the structure and location of files, we can include this and other critical information, that a peer or outside personnel should know concerning the project, in the `README.md` file.
+This file can be auto-generated when a project is initialized in a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be added manually aftwerwards by creating and adding a text file with the name `README.md` in the root directory of the repository.
+
+To speed up these initial steps, we provide a pre-populated template of such a structured study repository based on the HARPER protocol[@wang2022] under <https://gitlab-scm.partners.org/drugepi/harper>.
 
@@
 
-### Stage and commit changes
+### Stage and commit changes {#sec-stage-and-commit-changes}
 
 As introduced before, Git allows researchers to keep track of changes by making snapshots of the repository every time the state of a project is saved.
-In Git terms, this fundamental step is referred to as `commit`.
-Generally it is up to the user which files should be committed, how the commit should be documented in a `commit message` and when it should be performed.
+In Git terms, this fundamental step is referred to as a `commit`.
+Generally it is up to the user when to commit, which files to commit and how the commit should be documented in a `commit message`.
 
-To determine which files should be selected (or `staged`) for a new commit, the `git add` command is used followed by the name(s) of the files that should be committed or a `.` to stage all changes present in the current repository.
+To determine which files should be selected (or `staged`) for a new commit, the `git add` command is used followed by the name(s) of the files that should be committed or a '`.'` to stage all changes present in the current repository.
 
@@
 # Examplary commit command
-git commit -m 'Applied inclusion and exclusion criteria to cohort query'
+git commit -m 'added title to plot illustrating propensity score overlap'
 ```
+
+Every time a user makes a commit to create a snapshot of the work, `Git` creates a unique hash (a 40-character string created by a **simple hashing algorithm** \[SHA\]) which can be seen as an "ID" of the snapshot which enables user to comprehend every change made and revert back to any snapshot of the repository.
+@fig-stage-commit-RStudio-gui illustrates the equivalent step by using the RStudio GUI.
 
@@
 
-@fig-stage-commit-RStudio-gui illustrates the equivalent steps in using the RStudio GUI.
-
 ## Synchronize changes with remote repository
 
-### Local changes to remote (`push`)
+### Synchronize local changes to remote (`push`)
 
-To synchronize the local changes to our files, the next step involves synchronizing our commits with the remote repository on GitHub via the `push` command.
+To synchronize the local changes made to our files, the next step involves synchronizing our commits with the remote repository on GitHub via the `push` command.
 
@@
 
-In case we initialized the local repository as described in section @sec-gui-initialization-and-git-clone, a remote was already automatically configured. This can be checked via the following command.
+In case we initialized the local repository as described in section @sec-gui-initialization-and-git-clone, a remote was automatically configured. This can be checked via the following command.
 
@@
 
-This step is not needed if the intention of a researcher is just to keep track of local changes and there is no intention to share code, collaborate with others or work across different systems.
-However, most of the times researchers need to collaborate on a project and `git` can improve collaborative work significantly (@fig-collaborate).
+This step is not needed if the intention of a researcher is just to keep track of local changes and there is no wish to share code, collaborate with others or work across different systems.
+However, most of the times researchers intentionally use `Git` to collaborate on a project and remote repositories such as GitHub as this can improve collaborative work significantly (@fig-collaborate).
 
-### Remote to local
+### Synchronize remote to local (`pull`)
 
@@
 Occasionally it can occur that two collaborators made conflicting changes to the same file or even to the same line of code.
-While Git is generally good at automatically integrating (`merging`) new changes, Git will notify the user to resolve a conflict by editing the conflicting file if Git cannot automatically determine what is correct.
+While Git is generally good at automatically integrating (`merging`) new changes, Git will notify the user to resolve a major conflict by editing the conflicting file if Git cannot automatically determine what is correct ((e.g., when one collaborator has made changes to a line of code and the other completely deleted it).
 
@@
 
+One of the main motivations to adhere to this workflow is that in return we get rewarded with a detailed audit trail of changes made to our files.
+Probably, the best way to visualize and track changes is by browsing the `commit history` on the remote repository.
+@fig-GitHub-gui_changes illustrates this for the example commit we have previously performed in @sec-stage-and-commit-changes.
+A similar view for uncommitted changes is also provided in the RStudio GUI as displayed in @fig-stage-commit-RStudio-gui.
+
+For completeness, tracking uncommitted changes is also possible via the command line, but usually provides the user with a less intuitive output compared to what is visually provided by RStudio and GitHub.
+
+```{bash}
+git diff
+```
+
+A complete summary of committed changes with details on SHA hash, author, email, time and commit messages can be accessed via
+
+```{bash}
+git log
+```
+
 ## Keeping track of R software versions and paths
+
+A major advantage of modern programming languages, is that they provide a large pool of open-source packages.
+Packages are compilations of functions that automate frequently used procedures (e.g., querying new user cohorts[@tidyverse] or propensity score matching[@MatchIt]) which expedites the implementation of RWE studies, mitigates copy-paste errors and makes code much more friendly to read and review.
+
+A further step to improve reproducibility when using the R programming language is to ensure that anyone intending to reproduce analytic results from a `Git` repository is able to work off the same R package versions and dependencies without having to perform manual installations or path adjustments.
+These requirements are nothing exclusive to R, but apply to any type of programming language.
+
+### Reproducible R computing environments
+
+One way to achieve this is through self-contained computing environments via Docker images which are isolated and standardized environments which include everything needed to run and deploy applications and code regardless of a user's local operating system.[@boettiger2017a; @nüst2020] Another more lightweight, yet effective way is by using the `renv` R package[@renv], which helps creating **re**producible **env**ironments and manages R package dependencies for a given project.
+Through a lockfile, the `renv` package keeps track of the underlying version of R that is being used as well as all R packages, their versions and the installation source (e.g., <https://cran.r-project.org/>).
+To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, all that needs to be done is running the following command in R.
+
+```{r}
+renv::restore()
+```
+
+Appending the session info to the end of an R script can provide further context.
+
+```{r}
+sessionInfo()
+```
+
+### Absolute versus relative paths
+
+Another common issue when trying to reproduce analytic code is the use of absolute paths such as in the following example, where we try to call an `icd_code` file from the project's `protocol` sub-directory.
+
+```{r}
+icd_codes <- read.csv("C:\Users\Username\Projects\rwe_study\protocol\icd_codes.csv")
+```
+
+The use of absolute paths will make it impossible for anyone to run the code without having to manually adapt every single path throughout a repository which is tedious and time consuming.
+A better solution to this is the use of relative paths.
+A popular R package to implement this is the `here` package which enables file referencing in project-oriented workflows.
+The main function "`here"` creates paths relative to the top-level of the project directory (in our example `rwe_study`).
+Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any user re-running this script.
+
+```{r}
+library(here)
+icd_codes <- read.csv(here("data", "icd_codes.csv"))
+```
 
@@
 Also other GUIs to interact with Git...
-
-Docker
 
@@
 
-# Tables
-
-Create a table with most important git commands
-
-# Figures
+# Figures {.unnumbered}
 
@@
 
+```{r}
+{r}
+#| label: fig-GitHub-gui_changes
+#| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures.
+This commit history illustrates details about the commit (e.g., commit SHA hash and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
+#| out-width: 105%
+
+include_graphics(here("figures", "SFigure_4_commit_Github_gui.png"))
+```
+
 # References {.unnumbered}

### 2023-06-29 manuscript/manuscript.qmd
started writing up discussion

@@
 
+[**Short running title**]{.underline}: FAIRification of pharmacoepidemiological research
+
 [**Keywords:**]{.underline} Transparency, Reproducibility, Git, R, Version control, Real-world evidence
@@
 
-**Objective**
+**Purpose**
 
@@
 Transparency and reproducibility in conducting healthcare database studies in pharmacoepidemiology are critical scientific requirements for meaningful research.
-While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], the transparency around the actual implementation and analytical steps is still significantly lacking, especially when analyses are implemented via line programming steps in statistical programming languages like R[@RCore2023] or Python.
+While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], the transparency around the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line programming steps in statistical programming languages like R[@RCore2023] or Python.
 Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018] or health informatics[@bakken2019], the sharing of programming code is rather the rule than the exception, this practice is not as common in the (pharmaco-)epidemiological community.
@@
 
-Also other GUIs to interact with Git...
+In this practical tutorial, we introduced basic concepts of VCS for transparent and reproducible workflows and gave technical step-by-step instructions for the implementation using `Git` and R.
+We discussed frequently occurring challenges for reproducibility and how these can be overcome.
 
-Advantages of distributed VCS (Git):
+`Git` is a powerful tool that perfectly integrates into workflows of quantitative disciplines which have to handle analytic code.
+It further provides features that make it convenient for project-oriented setups and to collaborate at scale.
+This tutorial aimed to give an introduction to essential concepts that make it useful for pharmacoepidemiology research, but is by no means comprehensive as `Git` provides a vast amount of additional features which were not addressed in this manuscript.
+To get familiar with more advanced `Git` usage, we compiled a non-exhaustive list of further readings in the **Supplementary Material**.
+
+`Git` was primarily designed to work with source code files and works best for plain, text-based documents such programming code, rich text format (RTF), Markdown or LaTex files.
+There are limitations when it comes to tracking binary files like Word (.docx) and Excel (.xlsx) as these need to be interpreted by a program or a hardware processor[@blischak2016].
+While it can be still useful to track versions of these file types, it is important to recognize that modifications in such documents can't be easily inspected using Rstudio or GitHub.
+
+Moreover, when working with remote repositories, it is of utmost importance to ensure that sensitive information (such as passwords or patient-level data) are not accidentally shared.
+To prevent this, we recommend specifying such files in the `.gitignore` file before making the first commit or, more preferably, not storing them in the project directory in the first place.
+
+Despite its potential initial complexities to get acquanited with the `Git` workflow, distributed VCS and code sharing provide unique advantages:
 
@@
 # Figures {.unnumbered}
-
-```{r}
-#| label: Fig1
-#| fig-cap: "Clone a remote repository using the RStudio graphical user interface."
-#| eval: false
-#| echo: false
-
-fig1a <- ggdraw() +
-  draw_image(image_read(here("figures", "Figure_1a_RStudio.png")))
-
-fig1b <- ggdraw() +
-  draw_image(image_read(here("figures", "Figure_1b_RStudio.png")))
-
-fig1c <- ggdraw() +
-  draw_image(image_read(here("figures", "Figure_1c_RStudio.png")))
-
-fig1d <- ggdraw() +
-  draw_image(image_read(here("figures", "Figure_1d_RStudio.png")))
-
-figure_1 <- plot_grid(
-  fig1a, fig1b, fig1c, fig1d,
-  nrow = 2,
-  ncol = 2,
-  #labels = c("a)", "b)", "c)", "d)"),
-  hjust = 0.005,
-  label_size = 16
-  )
-
-png(
-  filename = here("figures", "Figure_1_RStudio_init.png"), 
-  width = 15, height = 10, units = "in", res=600
-  )
-figure_1
-invisible(dev.off())
-
-knitr::include_graphics(here::here("figures", "Figure_1_RStudio_init.png"))
-```
 
@@
 
-include_graphics(here("figures", "SFigure_4_commit_Github_gui.png"))
+include_graphics(here("figures", "Figure_6_commit_Github_gui.png"))
 ```

### 2023-06-29 manuscript/manuscript.qmd
first complete draft of manuscript and supplement (before co-author review)

@@
 ---
-title: "The FAIRification of pharmacoepidemiological research: A practical introduction to reproducible analytic workflows using git and R"
+title: "The FAIRification of research in real-world evidence: A practical introduction to reproducible analytic workflows using Git and R"
 
@@
 
-[**Word count:**]{.underline} xxxx words
+[**Word count:**]{.underline} 3,028 words
 
@@
 
-[**Supplementary material:**]{.underline}
+[**Supplementary material:**]{.underline} Supplementary Appendix and Supplementary Figures (pdf)
 
-[**Short running title**]{.underline}: FAIRification of pharmacoepidemiological research
+[**Short running title**]{.underline}: FAIRification of research in real-world evidence
 
@@
 
-[**Acknowledgments:**]{.underline} The authors would like to thank ...
+[**Acknowledgments:**]{.underline} None.
 
@@
 
-**Purpose**
-
-**Study Design and Settings**
-
-**Results**
-
-**Conclusions**
+Transparency and reproducibility are major prerequisites for meaningful real-world evidence (RWE) studies.
+Many advances have been made in the documentation and reporting of study protocols and results, but the version control and sharing of analytic code in RWE is not as established yet as in other quantitative disciplines like computational biology and health informatics.
+In this practical tutorial, we aim to give an introduction to distributed version control systems (VCS) tailored towards the FAIR (**F**indable, **A**ccessible, **I**nteroperable and **R**eproducible) implementation of RWE studies.
+To ease adoption, we provide detailed step-by-step instructions with practical examples on how the Git VCS and R programming language can be implemented into RWE workflows to facilitate reproducible analyses.
+We further discuss and showcase how these tools can be used to track changes, collaborate, disseminate and archive RWE studies through dedicated project repositories maintaining a complete audit trail of all relevant study documents.
 
@@
 
-Transparency and reproducibility in conducting healthcare database studies in pharmacoepidemiology are critical scientific requirements for meaningful research.
-While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], the transparency around the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line programming steps in statistical programming languages like R[@RCore2023] or Python.
-Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018] or health informatics[@bakken2019], the sharing of programming code is rather the rule than the exception, this practice is not as common in the (pharmaco-)epidemiological community.
+Transparency and reproducibility in conducting real-world evidence (RWE) studies are critical scientific requirements for meaningful research.
+While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], the transparency around the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023] or Python.
+Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community.
 
-Version control systems (VCS), such as *Git*, can thereby provide easy and powerful tools to keep track of the versioning of important files and documents such as protocols, analytical code, tables and figures.
+Version control systems (VCS), such as *Git*, can thereby provide easy and powerful tools to keep track of the versioning of important files and documents such as protocols, analytic code, tables and figures.
 Thereby, it can extend the principles of *FAIR* *data*[@garcía-closas2023] to the implementation of pharmacoepidemiological study workflows by making all components of a study **F**indable, **A**ccessible, **I**nteroperable, and **R**eproducible (**FAIR**).
 
-VCS enable researchers to track and resolve errors, collaborate with peers, and share resources instantaneously, for example, upon submission or publication of a manuscript.
-By design, VCS workflows automatically empower users to comment, track and compare changes made to files and hence, increase the ability to comprehend the evolution of a project over time while maintaining a complete audit trail of all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as "*final_analysis_1_rev_more_changes.R*" which is bad practice as this is highly error prone, lacking transparency for collaborating project members and outside personnel, and changes are difficult to track and reconcile.
-Particularly in disciplines like pharmacoepidemiology, where analyses can have far reaching impact on healthcare decisions, analytical code needs to be accessible and reproducible by anyone, not only the analyst who originally carried out the analysis.
+VCS enable researchers to track and resolve errors, collaborate with peers, and share resources instantaneously, for example, upon publication of a manuscript.
+By design, VCS workflows automatically empower users to comment, track and compare changes made to files and hence, increase the ability to comprehend the evolution of a project over time while maintaining a complete audit trail of all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as "*final_analysis_1_rev_more_changes.R*" which is bad practice as this is highly error prone, lacking transparency for collaborators, and changes are difficult to track and reconcile.
+Particularly in disciplines like pharmacoepidemiology, where analyses can have far reaching impact on healthcare decisions, analytic code needs to be accessible and reproducible.
 
-To increase the adoption of transparent and reproducible workflows in real-world evidence (RWE) studies using VCS, this tutorial aims to give a practical introduction tailored for pharmacoepidemiologists on how to set up, structure, and implement workflows using *Git*, which is the most frequently used VCS to date.[@blischak2016] We will first give a brief introduction to Git and how it can be used for collaboration and dissemination of study results using project repositories.
-Then we will provide a technical step-by-step guidance on how to integrate Git in an analytic workflow.
+To increase the adoption of transparent and reproducible workflows in RWE studies using VCS, this tutorial aims to give a practical introduction for pharmacoepidemiologists on how to set up, structure, and implement workflows using *Git*, which is the most popular VCS to date.[@blischak2016] We will first give a brief introduction to Git and its usage for collaboration and dissemination of study results through project repositories.
+Then we will provide a technical step-by-step guidance on how to integrate Git in analytic RWE workflows.
 We will additionally discuss and showcase important aspects of reproducibillity using the R open-source programming language, although the basic principles are applicable to any major coding language.
@@
 
-# A step-by-step introduction on how to use Git in a pharmacoepidemiology research study
+# A step-by-step introduction on how to use Git in a RWE study
 
@@
 
-To check if the installation succeeded and show the available version, the following prompt in the CLI can be used.
+To check if the installation succeeded, the following prompt in the CLI can be used, which will output the available version.
 
@@
 
-An alternative to creating a new repository locally is through creating a new project in the remote repository (in this case GitHub) and then creating a local copy, a process which is usually referred to as `cloning`.
+An alternative to creating a new repository locally is through initiating a new project in the remote repository (in this case GitHub) and then creating a local copy, a process which is usually referred to as `cloning`.
 
 To do so, users will need to log into their remote repository account and navigate to the section `New repository` (**Supplementary Figure 1**).
-This will take a user to a form to populate basic information of the new project such as the repository name (`rwe_study`) and choose basic setting such as if the repository should be private or public and if a `README.md` (see @sec-repository-structure) and `.gitignore` (see @sec-stage-and-commit-changes) file should be automatically created (**Supplementary Figure 2**).
+This will take a user to a form to populate basic information of the new project such as the repository name (`rwe_study`) and choose basic settings such as if the repository should be private or public and if `README.md` (see @sec-repository-structure) and `.gitignore` (see @sec-stage-and-commit-changes) files should be automatically created (**Supplementary Figure 2**).
 
@@
 
-At this step, we now can populate the repository and start working on study-relevant documents.
-To enhance clarity, it is recommended to use a logical structure and informative and transparent sub-directory and file names (a minimal example is illustrated in @fig-structure).
-To document the structure and location of files, we can include this and other critical information, that a peer or outside personnel should know concerning the project, in the `README.md` file.
-This file can be auto-generated when a project is initialized in a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be added manually aftwerwards by creating and adding a text file with the name `README.md` in the root directory of the repository.
+At this step, we now can populate the repository and start working on study-relevant documents and analyses.
+To enhance clarity, it is recommended to use a logical structure and an informative nomenclature for sub-directories and file names (a minimal example is illustrated in @fig-structure).
+To document the structure and location of files, we can include this, and other critical information about the project, in the `README.md` file.
+This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added aftwerwards by creating a text file with the name `README.md` in the root directory of the repository.
 
@@
 
-Every time a user makes a commit to create a snapshot of the work, `Git` creates a unique hash (a 40-character string created by a **simple hashing algorithm** \[SHA\]) which can be seen as an "ID" of the snapshot which enables user to comprehend every change made and revert back to any snapshot of the repository.
-@fig-stage-commit-RStudio-gui illustrates the equivalent step by using the RStudio GUI.
+Every time a user makes a commit to create a snapshot of the work, `Git` creates a unique hash (a 40-character string created by a **simple hashing algorithm** \[SHA\]) which can be seen as an "ID" of the snapshot which enables users to comprehend every change made and revert back to any snapshot of the repository.
+@fig-stage-commit-RStudio-gui illustrates the equivalent steps in RStudio GUI.
 
-If any file(s) in the study repository should not be tracked and hence not be synchronized with the remote repository (e.g., because they include confidential information that should not be visible to others like database credentials), these files can be specified in the `.gitignore` file which needs to be located in the root directory of the repository.
-To create a `.gitignore` file, either auto-generate it upon initialization of the repository on GitHub (**Supplementary Figure 2**) or manually add a text file called `.gitignore` to the root directory.
+If any file(s) in the study repository should not be tracked and hence not be synchronized with the remote repository (e.g., because they include confidential information that should not be visible to others like database credentials), these can be specified in the `.gitignore` file which is typically located in the root directory of the repository.
+To create a `.gitignore` file, a user can either auto-generate it upon initialization of the repository on GitHub (**Supplementary Figure 2**) or manually add a text file called `.gitignore` to the root directory.
 
@@
 
-To synchronize the local changes made to our files, the next step involves synchronizing our commits with the remote repository on GitHub via the `push` command.
+To synchronize the local changes, the next step involves the upload of our commits to the remote repository on GitHub via the `push` command.
 
@@
 
-If there is no remote established yet, this can be configured using the following commands.
+If there is no remote established yet, this can be retrospectively configured using the following commands.
 
 ```{bash}
-git remote add origin https://github.com/USERNAME/rwe_study.git
+git remote add origin https://github.com/USER/rwe_study.git
 git branch -M main
@@
 This step is not needed if the intention of a researcher is just to keep track of local changes and there is no wish to share code, collaborate with others or work across different systems.
-However, most of the times researchers intentionally use `Git` to collaborate on a project and remote repositories such as GitHub as this can improve collaborative work significantly (@fig-collaborate).
+However, most of the times researchers use `Git` for the purpose of collaborating on a project as depicted in @fig-collaborate.
 
@@
 Occasionally it can occur that two collaborators made conflicting changes to the same file or even to the same line of code.
-While Git is generally good at automatically integrating (`merging`) new changes, Git will notify the user to resolve a major conflict by editing the conflicting file if Git cannot automatically determine what is correct ((e.g., when one collaborator has made changes to a line of code and the other completely deleted it).
+While Git is generally good at automatically integrating (`merging`) new changes, Git will notify the user to resolve a major conflict by editing the conflicting file if Git cannot automatically determine what is correct (e.g., if one collaborator has made changes to a line of code and the other completely deleted it).
 
@@
 
-One of the main motivations to adhere to this workflow is that in return we get rewarded with a detailed audit trail of changes made to our files.
-Probably, the best way to visualize and track changes is by browsing the `commit history` on the remote repository.
+One of the main motivations to adhere to this workflow is that in return we get rewarded with a detailed audit trail of changes made to our study files.
+Probably the best way to visualize and track changes is by browsing the `commit history` on the remote repository.
 @fig-GitHub-gui_changes illustrates this for the example commit we have previously performed in @sec-stage-and-commit-changes.
-A similar view for uncommitted changes is also provided in the RStudio GUI as displayed in @fig-stage-commit-RStudio-gui.
+A similar view for uncommitted changes in the RStudio GUI is also provided in @fig-stage-commit-RStudio-gui.
 
@@
 
-A complete summary of committed changes with details on SHA hash, author, email, time and commit messages can be accessed via
+A complete summary of committed changes with details on the SHA hash, author, email, time and commit messages can be accessed via
 
@@
 ```
+
+While there are many more useful `Git` features such as `branching` (see **Supplementary Material**), the workflow of staging, committing and pushing/pulling (@fig-workflow) may already suffice for the majority of pharmacoepidemiologists.
 
@@
 
-A further step to improve reproducibility when using the R programming language is to ensure that anyone intending to reproduce analytic results from a `Git` repository is able to work off the same R package versions and dependencies without having to perform manual installations or path adjustments.
+A further step to improve reproducibility when using the R programming language is to ensure that anyone, intending to reproduce analytic results from a `Git` repository, is able to work off the same R package versions and dependencies without having to perform manual installations or path adjustments.
 These requirements are nothing exclusive to R, but apply to any type of programming language.
@@
 Through a lockfile, the `renv` package keeps track of the underlying version of R that is being used as well as all R packages, their versions and the installation source (e.g., <https://cran.r-project.org/>).
-To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, all that needs to be done is running the following command in R.
+To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, the only step to necessary is running the following command in R.
 
@@
 
-Appending the session info to the end of an R script can provide further context.
+Appending the session info to the end of an R script can provide further context about the dependencies and versions of an analytic script.
 
@@
 ```{r}
-icd_codes <- read.csv("C:\Users\Username\Projects\rwe_study\protocol\icd_codes.csv")
+icd_codes <- read.csv('C:/User/Projects/rwe_study/protocol/icd_codes.csv')
 ```
 
-The use of absolute paths will make it impossible for anyone to run the code without having to manually adapt every single path throughout a repository which is tedious and time consuming.
+The use of absolute paths will make it impossible for any collaborator to run the code without having to manually adapt every single path throughout a repository which is tedious and time consuming.
 A better solution to this is the use of relative paths.
 A popular R package to implement this is the `here` package which enables file referencing in project-oriented workflows.
-The main function "`here"` creates paths relative to the top-level of the project directory (in our example `rwe_study`).
+The main function "`here"` creates paths relative to the top-level of the project directory (in our example the `rwe_study` directory).
 Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any user re-running this script.
@@
 
-`Git` is a powerful tool that perfectly integrates into workflows of quantitative disciplines which have to handle analytic code.
-It further provides features that make it convenient for project-oriented setups and to collaborate at scale.
-This tutorial aimed to give an introduction to essential concepts that make it useful for pharmacoepidemiology research, but is by no means comprehensive as `Git` provides a vast amount of additional features which were not addressed in this manuscript.
-To get familiar with more advanced `Git` usage, we compiled a non-exhaustive list of further readings in the **Supplementary Material**.
+`Git` is a powerful tool that perfectly integrates into workflows of quantitative disciplines handling analytic code.
+It further provides features that make it convenient for project-oriented setups and collaboration at scale.
+This tutorial is by no means comprehensive as `Git` provides a vast amount of additional features which were not addressed in this manuscript.
+To get familiar with more advanced `Git` usage, we compiled a non-exhaustive list of further suggested readings and resources in the **Supplementary Material**.
 
-`Git` was primarily designed to work with source code files and works best for plain, text-based documents such programming code, rich text format (RTF), Markdown or LaTex files.
-There are limitations when it comes to tracking binary files like Word (.docx) and Excel (.xlsx) as these need to be interpreted by a program or a hardware processor[@blischak2016].
-While it can be still useful to track versions of these file types, it is important to recognize that modifications in such documents can't be easily inspected using Rstudio or GitHub.
+`Git` was primarily designed to work with source code files and works best for plain, text-based documents such as programming code, rich text format (RTF), Markdown or LaTex files.
+There are limitations when it comes to tracking binary files like Word (.docx) and Excel (.xlsx) as these need to be interpreted by a program or a hardware processor.[@blischak2016] While it can be still useful to track versions of these file types using `Git`, it is important to recognize that modifications in such documents can't be easily inspected using RStudio or GitHub.
 
@@
 
-Despite its potential initial complexities to get acquanited with the `Git` workflow, distributed VCS and code sharing provide unique advantages:
+Despite the potential initial complexities to get acquanited with the `Git` workflow, distributed VCS and code sharing provide unique advantages, such as a full audit trail of which changes were made by whom, when and why.
+Git makes it easier for users to go back to any snapshot and resolve errors more efficiently.
+Further, remote repositories are ideal platforms to deploy and archive scripts and documents, facilitate collaboration and make it very convenient to share code.
+Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book] which makes shared code interoperable even without direct access to the raw data.
+Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unneccesary redundancies through sharing of code libraries, functions and algorithms.
 
--   Full audit trail of which changes were made by whom, when and why
+{{< pagebreak >}}
 
--   Even if the code is not shared, Git makes it easier for an individual to go back to the code
+# References {.unnumbered}
 
--   Git can serve as internal and external repository for code and made accessible to everyone, e.g. along with manuscript without copy-paste into word files or similar
-
--   It is free and open-source and reliable
-
--   Distributed VCS fully mirror entire repositories and hence every local clone is a full backup of the data
+::: {#refs}
+:::
 
@@
 #| label: fig-RStudio-gui
-#| fig-cap: "Clone a remote repository using the RStudio graphical user interface."
+#| fig-cap: "Steps to clone a remote repository using the RStudio graphical user interface."
 #| eval: true
@@
 ```{r}
-{r}
 #| label: fig-GitHub-gui_changes
@@
 This commit history illustrates details about the commit (e.g., commit SHA hash and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
+#| eval: true
+#| echo: false
 #| out-width: 105%
 
-include_graphics(here("figures", "Figure_6_commit_Github_gui.png"))
+knitr::include_graphics(here("figures", "Figure_6_commit_Github_gui.png"))
 ```
-
-# References {.unnumbered}

### 2023-06-29 manuscript/supplement.qmd
first complete draft of manuscript and supplement (before co-author review)

@@
 title: "Supplementary Appendix"
-subtitle: "The FAIRification of pharmacoepidemiological research: A practical introduction to reproducible analytical workflows using git and R"
+subtitle: "The FAIRification of research in real-world evidence: A practical introduction to reproducible analytic workflows using Git and R"
 code-fold: true
@@
     crossref:
-      fig-title: "**Supplementary Figure**"
+      fig-title: "Supplementary Figure"
     toc: true
@@
 
+# Supplementary material
+
+## Branching
+
+Sometimes it is necessary to implement changes to a document or code without wanting to interfere with a current \"`production`\" version of a document or code.
+However, a researcher may still want to be able to \"`merge`\" back potential changes, e.g.
+to integrate new analyses into a script or to develop a new set of features or functions to an application.
+To do this in a safe, transparent and efficient manner, `Git` provides the option to **branch** our repository.
+The default branch of any repository is the `main` branch.
+
+Branching is a concept that is highly adapted in software development where the `main` branch reflects the software version that is deployed for public use (i.e., the production version) and a `development` branch that is the version that implements new features (i.e., the beta version).
+Hence, `branches` are used in order to have the flexibility to develop new features without jeopardizing the current `main` and `development` versions.
+
+Branching can have two very useful use cases for adaption in real-world evidence (RWE) studies.
+First, new analyses or improvements of existing analyses can be implemented without having to "mess" with existing code which works perfectly.
+Once a new analysis or visualization of results was established in a new `named branch` and works without errors, this branch can be `merged` back to the `main` branch (as depicted in the illustration below).
+
+```{r}
+#| label: branching
+#| out-width: 100%
+
+include_graphics(here("figures", "SFigure_x_branching.png"))
+```
+
+Second, branches can be re-purposed as backup versions of the repository at project landmarks.
+Examples could range from the first round of review among co-authors to the initial submission of a manuscript draft to a journal, corresponding rounds of revisions and the final version.
+This way it can be ensured that also prior versions of analyses can be reproduced without problems.
+
+To create a new branch and switch to it from the main branch, you can run the `git checkout` command with the -b switch. Let's say we call the new branch `lasso_variable_selection`.
+
+```{bash, eval=FALSE, echo=TRUE}
+git checkout -b lasso_variable_selection
+```
+
+This is shorthand for:
+
+```{bash, eval=FALSE, echo=TRUE}
+git branch lasso_variable_selection
+git checkout lasso_variable_selection
+```
+
+It is possible to switch between multiple branches back and forth and it's important to keep in mind that changes need to be committed before switch to the new branch branch.
+A user can e.g.
+switch back to the master branch by running the following command:
+
+```{bash, eval=FALSE, echo=TRUE}
+git checkout master
+```
+
+Suppose we have decided that the `lasso_variable_selection` approach led to a better balancing of the two new user cohorts after propensity score matching and we are ready to merge the implemented changes back into our main branch.
+This can be achieved by checking out the branch you wish to merge into (in this case `main` and then run the git merge command:
+
+```{bash, eval=FALSE, echo=TRUE}
+git checkout main
+git merge lasso_variable_selection
+```
+
+If the branch is merged and there is no desire anymore to use this branch any longer, the `lasso_variable_selection` branch can be deleted with the following command:
+
+```{bash, eval=FALSE, echo=TRUE}
+git branch -d lasso_variable_selection
+```
+
+These steps can also easily be done on remote repositories like GitHub or GitLab.
+The only difference to keep in mind is that GitHub counterintuitively calls the procedure of merging two branches `pull request` while GitLab refers to it as `merge request`.
+
+In case `Git` recognizes any clashes or incompatibilities of the new code with the existing code it will notify the user and give us the chance to reconcile inconsitencies.
+
+{{< pagebreak >}}
+
+## Advanced readings on `Git` and other resources
+
+-   Main `Git` website to download and learn more about `Git:` <https://git-scm.com>
+
+-   [Pro Git](https://git-scm.com/book/en/v2) by Scott Chacon and Ben Straub.
+
+-   Cheatsheet on most important `Git` commands: <https://education.github.com/git-cheat-sheet-education.pdf>
+
+-   GitHub Desktop App: <https://desktop.github.com/>
+
+-   Git manual by *Atlassian* (provider of the `Bitbucket` platform): <https://www.atlassian.com/git>
+
+-   Coursera courses:
+
+    -   [Introduction to Git and GitHub](https://www.coursera.org/learn/introduction-git-github) (Google)
+
+    -   [Version Control with Git](https://www.coursera.org/learn/version-control-with-git) (Atlassian)
+
+    -   [Getting Started with Git and GitHub](https://www.coursera.org/learn/getting-started-with-git-and-github) (IBM)
+
+    -   [Making Data Science Work for Clinical Reporting](https://www.coursera.org/learn/making-data-science-work-for-clinical-reporting?=) (Genentech)
+
+-   Quarto (open-source scientific and technical publishing system for reproducible research): <https://quarto.org>
+
+-   Choosing a suitable license for a remote repository: Public repositories need to be licensed to be truly open source.
+If no license is provided, the work is by default under exclusive copyright.
+The website <https://choosealicense.com> provides an excellent overview of the different types of licenses.
+
+-   Useful Git repositories for observational healthcare research:
+
+    -   Division of Pharmacoepidemiology and Pharmacoeconomics: [https://gitlab-scm.partners.org/drugepi](https://gitlab-scm.partners.org/drugepi/harper){.uri}
+
+    -   Observational Health Data Sciences and Informatics (OHDSI): <https://github.com/OHDSI>
+
+    -   Darwin EU® Private Development Repository: <https://github.com/darwin-eu-dev>
+
+    -   Pharmaverse (Network of developers of curated open source R packages for clinical reporting): [https://pharmaverse.org](https://pharmaverse.orghttps://pharmaverse.org) & <https://github.com/pharmaverse/>
+
+{{< pagebreak >}}
+
 # Supplementary Figures
@@
 
-include_graphics(here("manuscript/figures", "SFigure_1a_new_repo.png"))
+include_graphics(here("figures", "SFigure_1a_new_repo.png"))
 ```
@@
 
-include_graphics(here("manuscript/figures", "SFigure_2_repo_info.png"))
+include_graphics(here("figures", "SFigure_2_repo_info.png"))
 ```
@@
 #| fig-cap: "Clone a GitHub repository using the HTTPS link and the git clone command"
-#| out-width: 115%
-
-include_graphics(here("manuscript/figures", "SFigure_3_clone.png"))
+#| out-width: 105%
+include_graphics(here("figures", "SFigure_3_clone.png"))
 ```
@@
 
-# References {.unnumbered}
+```{r}
+#| label: fig-GitHub-gui_changes
+#| eval: false
+#| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures.
+This commit history illustrates details about the commit (e.g., commit SHA hash and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
+#| out-width: 105%
+
+include_graphics(here("figures", "SFigure_4_commit_Github_gui.png"))
+```

### 2023-06-30 manuscript/manuscript.qmd
corrected word count
added missing citation for the here package
corrected gitlab to github url

@@
 
-[**Word count:**]{.underline} 3,028 words
+[**Word count:**]{.underline} 3,033 words
 
@@
 cd ~/rwe_study
-git clone  https://gitlab.com/USERNAME/rwe_study.gitlab.io.git
+git clone https://github.com/USER/rwe_study.git
 ```
@@
 A better solution to this is the use of relative paths.
-A popular R package to implement this is the `here` package which enables file referencing in project-oriented workflows.
+A popular R package to implement this is the `here` package[@here] which enables file referencing in project-oriented workflows.
 The main function "`here"` creates paths relative to the top-level of the project directory (in our example the `rwe_study` directory).

### 2023-06-30 manuscript/manuscript.qmd
included another useful reference

@@
 While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], the transparency around the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023] or Python.
-Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community.
+Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018;@perez-riverol2016] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community.
 

### 2023-07-03 manuscript/supplement.qmd
started adding an end-to-end example

@@
 
+## Example for an end-to-end workflow using Git
+
+This section aims to summarize a minimally commented end-to-end workflow of incorporating and tracking changes for a an analysis step of a RWE study.
+To that end, we use the same example as in the main manuscript.
+
+Step 1: In case we work with multiple branches (see @sec-branching below), we can first check if we are on the correct branch where we want to implement our changes.
+
+```{bash, eval = FALSE}
+# check the current branch
+git branch
+```
+
+Step 2: Before initiating any new changes, it can be a good idea to check if a collaborator, who may work on the same branch, made any changes.
+
+```{bash, eval = FALSE}
+# git pull changes
+git pull
+```
+
+Step 3: We make changes to the analysis script \`scripts/03_propensity_score_analysis.R\` by adding a title to a plot illustrating the propensity score overlap.
+
+```{r, eval = FALSE}
+# load data
+library(smdi)
+library(gtsummary)
+library(tidyverse)
+library(here)
+library(smdi)
+
+# we use  data from the smdi R package
+data <- smdi::smdi_data_complete
+
+# step 1: compute propensity of exposure assignment
+
+# 1.1 select variables to include in propensity 
+ps_covars <- data %>% 
+  dplyr::select(-exposure) %>% 
+  names()
+
+# 1.2: define formula for the GLM
+exposure_form <- as.formula(paste("exposure ~ ", paste(ps_covars, collapse = " + ")))
+
+# 1.3: compute odds
+exposure_fit <- stats::glm(
+  exposure_form,
+  data = data,
+  family = "binomial"
+  )
+
+# 1.4: show odds
+odds_table <- exposure_fit %>% 
+  gtsummary::tbl_regression(exponentiate = T)
+
+# step 2: compute propensity scores and visualize overlap before matching
+exposure_plot <- data %>% 
+  dplyr::mutate(ps = fitted(exposure_fit))
+
+# step 3: plot density/overlap before matching
+ps_plot <- exposure_plot %>% 
+  ggplot2::ggplot(ggplot2::aes(x = ps, fill = factor(exposure))) +
+  ggplot2::geom_density(alpha = .5) +
+  ggplot2::theme_bw() +
+  ggplot2::labs(
+    title = "Propensity score overlap",
+    x = "Pr(exposure)",
+    y = "Density",
+    fill = "Exposed"
+    )
+
+png(filename = here("figures", "Figure_1_propensity_score_overlap.png"), width = 15, height = 10, units = "in", res=600)
+ps_plot + ggplot2::theme(plot.title=element_text(size=70))
+dev.off()
+```
+
 ## Branching
 
-Sometimes it is necessary to implement changes to a document or code without wanting to interfere with a current \"`production`\" version of a document or code.
-However, a researcher may still want to be able to \"`merge`\" back potential changes, e.g.
+Sometimes it is necessary to implement changes to a document or code without wanting to interfere with a current "`production`" version of a document or code.
+However, a researcher may still want to be able to "`merge`" back potential changes, e.g.
 to integrate new analyses into a script or to develop a new set of features or functions to an application.

### 2023-07-03 manuscript/manuscript.qmd
incorporated Shirley's edits and comments

--- 2023-06-30 manuscript/manuscript.qmd
@@
 
-[**Word count:**]{.underline} 3,033 words
+[**Word count:**]{.underline} 3,239 words
 
@@
 
-[**Disclosures/COI:**]{.underline} Janick Weberpals reports no conflicts of interest.
+[**Disclosures/COI:**]{.underline} Janick Weberpals and Shirley Wang report no conflicts of interest.
 
-[Analytical code sharing statement:]{.underline} This manuscript was written using `Quarto` version 1.3.353 (<https://quarto.org/>) and R version 4.1.2.
+[Analytical code sharing statement:]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
 All materials can be found at <https://gitlab-scm.partners.org/drugepi/fair-epi>.
@@
 
-Transparency and reproducibility are major prerequisites for meaningful real-world evidence (RWE) studies.
-Many advances have been made in the documentation and reporting of study protocols and results, but the version control and sharing of analytic code in RWE is not as established yet as in other quantitative disciplines like computational biology and health informatics.
+Transparency and reproducibility are major prerequisites for conducting meaningful real-world evidence (RWE) studies that are fit for decision-making.
+Many advances have been made in the documentation and reporting of study protocols and results, but the principles for version control and sharing of analytic code in RWE are not yet as established as in other quantitative disciplines like computational biology and health informatics.
 In this practical tutorial, we aim to give an introduction to distributed version control systems (VCS) tailored towards the FAIR (**F**indable, **A**ccessible, **I**nteroperable and **R**eproducible) implementation of RWE studies.
-To ease adoption, we provide detailed step-by-step instructions with practical examples on how the Git VCS and R programming language can be implemented into RWE workflows to facilitate reproducible analyses.
-We further discuss and showcase how these tools can be used to track changes, collaborate, disseminate and archive RWE studies through dedicated project repositories maintaining a complete audit trail of all relevant study documents.
+To ease adoption, we provide detailed step-by-step instructions with practical examples on how the Git VCS and R programming language can be implemented into RWE study workflows to facilitate reproducible analyses.
+We further discuss and showcase how these tools can be used to track changes, collaborate, disseminate and archive RWE studies through dedicated project repositories that maintain a complete audit trail of all relevant study documents.
 
@@
 
-Transparency and reproducibility in conducting real-world evidence (RWE) studies are critical scientific requirements for meaningful research.
-While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], the transparency around the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023] or Python.
-Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community.
+Real-world evidence (RWE) studies that make secondary use of routinely collected health data captured in electronic health records and claims are increasingly being used to inform drug development, regulatory and coverage decisions, as well as clinical practice.[@fdaRWE2018; @eichler2015; @cioms2023; @makady2017] Considering the impact that RWE studies can have on patient healthcare, transparent and reproducible conduct of RWE studies is critically important.
+While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], guidance on transparency regarding the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023], SAS or Python.
+Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018; @perez-riverol2016] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community.
+As open source tools like R packages and Shiny dashboards[@pharmaverse2023] become increasingly used for for clinical trial reporting and regulatory submissions[@positPharma], the adoption of version control and code sharing practices is a crucial and timely topic for the field of RWE.
 
-Version control systems (VCS), such as *Git*, can thereby provide easy and powerful tools to keep track of the versioning of important files and documents such as protocols, analytic code, tables and figures.
-Thereby, it can extend the principles of *FAIR* *data*[@garcía-closas2023] to the implementation of pharmacoepidemiological study workflows by making all components of a study **F**indable, **A**ccessible, **I**nteroperable, and **R**eproducible (**FAIR**).
+Version control systems (VCS), such as *Git*, provide powerful tools to keep track of the versioning of important files and documents such as protocols, analytic code, tables and figures.
+Thereby, VCS can extend the principles of *FAIR* *data*[@garcía-closas2023] to the implementation of pharmacoepidemiological study workflows by making all components of a study **F**indable, **A**ccessible, **I**nteroperable, and **R**eproducible (**FAIR**).
 
 VCS enable researchers to track and resolve errors, collaborate with peers, and share resources instantaneously, for example, upon publication of a manuscript.
-By design, VCS workflows automatically empower users to comment, track and compare changes made to files and hence, increase the ability to comprehend the evolution of a project over time while maintaining a complete audit trail of all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as "*final_analysis_1_rev_more_changes.R*" which is bad practice as this is highly error prone, lacking transparency for collaborators, and changes are difficult to track and reconcile.
-Particularly in disciplines like pharmacoepidemiology, where analyses can have far reaching impact on healthcare decisions, analytic code needs to be accessible and reproducible.
+By design, VCS workflows automatically empower users to comment, track and compare changes made to files and hence, increase the ability to comprehend the evolution of a project over time while maintaining a complete audit trail of changes to all documents.[@ram2013] Without VCS, this is often naturally done by assigning uninformative file names such as "*final_analysis_1_rev_more_changes.R*" which is bad practice as this is highly error prone, lacking transparency for collaborators, and changes are difficult to track and reconcile.
+Particularly in disciplines like pharmacoepidemiology, where analyses can have far reaching impact on healthcare policy and practice decisions, analytic code needs to be accessible and reproducible.
 
@@
 Then we will provide a technical step-by-step guidance on how to integrate Git in analytic RWE workflows.
-We will additionally discuss and showcase important aspects of reproducibillity using the R open-source programming language, although the basic principles are applicable to any major coding language.
+We will additionally discuss and showcase important aspects of reproducibility using the R open-source programming language, although the basic principles are applicable to any major coding language.
 
@@
 
-These platforms make it possible to synchronize local Git repositories across multiple project members and thereby improve collaboration.
+These remote platforms make it possible to synchronize local Git repositories across multiple project members and thereby improve collaboration.
 They further complement Git functionalities by providing a graphical user interface (GUI) to visualize changes made to files and often offer advanced features such as the automation of workflows using continuous integration and deployment (CI/CD) or, most recently, AI-powered virtual coding assistants (e.g., GitHub co-pilot).
@@
 In this section, we want to give a technical introduction on how Git and remote repositories can be used in RWE studies.
-For the following examples, we focus on workflows utilizing R/RStudio ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository providers, respectively.
+For the following examples, we focus on workflows utilizing R/RStudio ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as a remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository providers, respectively.
 We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated GUI.
@@
 ```{bash}
+# check installed git version
 git --version
@@
 
-## Configuration and first time use
+## Configuration and first-time use
 
@@
 ```{bash}
+# global configuration of name and email address
 git config --global user.name "First Last"
@@
 ```{bash}
+# create and navigate to study directory
 mkdir ~/rwe_study
@@
 ```{bash}
+# initialize git tracking
 git init
@@
 ```{bash}
+# create and navigate to study directory and clone 
+# the git repository which was already created on GitHub
 mkdir ~/rwe_study
@@
 
-To speed up these initial steps, we provide a pre-populated template of such a structured study repository based on the HARPER protocol[@wang2022] under <https://gitlab-scm.partners.org/drugepi/harper>.
+To speed up these initial steps, we provide a pre-populated template of such a structured study repository based on the HARPER protocol[@wang2022] template under <https://gitlab-scm.partners.org/drugepi/harper>.
+HARPER is a harmonized study protocol template endorsed by the International Society of Pharmacoepidemiology and the International Society of Pharmacoeconomics and Outcomes Research.
+The pre-populated structured study repository template is publicly available and can be `forked`[@GitFork] to initiate a new project (see instructions in the template repository).
 
@@
 
-The most essential steps of a git workflow are summarized in @fig-workflow.
+The most essential steps of a git workflow are summarized in @fig-workflow and in the **Supplementary Material**.
 
@@
 In Git terms, this fundamental step is referred to as a `commit`.
-Generally it is up to the user when to commit, which files to commit and how the commit should be documented in a `commit message`.
+Generally, it is up to the user when to commit, which files to commit and how the commit should be documented in a `commit message`.
 
@@
 ```{bash}
+# push local changes to remote repository
 git push
@@
 ```{bash}
+# check remote configuration
 git remote -v
@@
 ```{bash}
+# add a new remote repository connection
 git remote add origin https://github.com/USER/rwe_study.git
@@
 ```{bash}
+# pull remote repository version of code to local
 git pull
@@
 ```{bash}
+# display differences between uncommitted changes
 git diff
@@
 ```{bash}
+# display commit history
 git log
@@
 
-A major advantage of modern programming languages, is that they provide a large pool of open-source packages.
+A major advantage of modern programming languages is that they provide a large pool of open-source packages.
 Packages are compilations of functions that automate frequently used procedures (e.g., querying new user cohorts[@tidyverse] or propensity score matching[@MatchIt]) which expedites the implementation of RWE studies, mitigates copy-paste errors and makes code much more friendly to read and review.
@@
 It further provides features that make it convenient for project-oriented setups and collaboration at scale.
-This tutorial is by no means comprehensive as `Git` provides a vast amount of additional features which were not addressed in this manuscript.
+This tutorial is by no means comprehensive as `Git` provides a vast number of additional features which were not addressed in this manuscript.
 To get familiar with more advanced `Git` usage, we compiled a non-exhaustive list of further suggested readings and resources in the **Supplementary Material**.
@@
 
-Despite the potential initial complexities to get acquanited with the `Git` workflow, distributed VCS and code sharing provide unique advantages, such as a full audit trail of which changes were made by whom, when and why.
+Despite the potential initial complexities to get acquainted with the `Git` workflow, distributed VCS and code sharing provide unique advantages, such as a full audit trail of which changes were made by whom, when and why.
 Git makes it easier for users to go back to any snapshot and resolve errors more efficiently.
 Further, remote repositories are ideal platforms to deploy and archive scripts and documents, facilitate collaboration and make it very convenient to share code.
-Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book] which makes shared code interoperable even without direct access to the raw data.
-Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unneccesary redundancies through sharing of code libraries, functions and algorithms.
+Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book], Sentinel[@brown2020] and PCORnet[@fleurence2014], which makes shared code interoperable even without direct access to the same data source.
+Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unnecessary redundancies through sharing of code libraries, functions and algorithms.
+
+Regulators[@fdaRWE2021], funding agencies[@NIH2023;@PCORI2023] and HTA bodies[@kent2021] have increasing expectations regarding the provenance, sharing and audit trial of study documents including analytic code as a prerequisite for impactful and credible RWE studies.
+This manuscript addresses this timely topic and will hopefully motivate and enable new and experienced pharmacoepidemiologists to integrate VCS workflows and code sharing into their daily work.
 
@@
 #| label: fig-structure
-#| fig-cap: "Minimal example of a transparent repository structure containing relevant study documents"
+#| fig-cap: "Minimal example of a transparent repository structure containing relevant study documents.
+Abbreviations: .git = Git sub-directory which includes all files to keep track of changes (created as a result of git init command), renv = R environment sub-directory which contains the project library and other project-specific files and settings needed to manage R package versions and the R environment, renv.lock = R environment lockfile, describing the R version and R package versions used in a project."
 #| eval: true
@@
 #| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures.
-This commit history illustrates details about the commit (e.g., commit SHA hash and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
+This commit history illustrates details about the commit (e.g., commit SHA [simple hasing algorithm] and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
 #| eval: true

### 2023-07-03 manuscript/supplement.qmd
incorporated Shirley's edits and comments

--- 2023-06-29 manuscript/supplement.qmd
@@
 subtitle: "The FAIRification of research in real-world evidence: A practical introduction to reproducible analytic workflows using Git and R"
-code-fold: true
 echo: false
@@
     toc-depth: 2
-    number-sections: false
+    number-sections: true
     colorlinks: true
@@
 
-## Branching
+## Example for an end-to-end workflow using Git
 
-Sometimes it is necessary to implement changes to a document or code without wanting to interfere with a current \"`production`\" version of a document or code.
-However, a researcher may still want to be able to \"`merge`\" back potential changes, e.g.
+This section aims to summarize a minimally commented end-to-end workflow for incorporating and tracking changes for an analysis step of an exemplary RWE study.
+To that end, we use the same example as in the main manuscript.
+
+Step 1: In case we work with multiple branches (see @sec-branching below), we can first check if we are on the correct branch where we want to implement our changes.
+
+```{bash, eval = FALSE, echo = TRUE}
+# check the current branch
+git branch
+```
+
+Step 2: Before initiating any new changes, it can be a good idea to check if a collaborator, who may work on the same branch, made any changes.
+
+```{bash, eval = FALSE, echo = TRUE}
+# git pull changes
+git pull
+```
+
+Step 3: We make changes to the analysis script `scripts/03_propensity_score_analysis.R` by adding a title to a plot illustrating the propensity score overlap.
+
+```{r, eval = FALSE, echo = TRUE}
+# load libraries
+library(tidyverse)
+
+# [...]
+
+# add a title to propensity score plot
+ps_plot <- exposure_plot %>% 
+  ggplot2::ggplot(ggplot2::aes(x = ps, fill = factor(exposure))) +
+  ggplot2::geom_density(alpha = .5) +
+  ggplot2::theme_bw() +
+  ggplot2::labs(
+    # added the title here
+    title = "Propensity score overlap",
+    x = "Pr(exposure)",
+    y = "Density",
+    fill = "Exposed"
+    )
+
+ps_plot
+
+# [...]
+```
+
+Step 4: After having all changes made, we stage the files we want to commit next.
+
+```{bash, eval = FALSE, echo = TRUE}
+# stage command
+git add 'scripts/03_propensity_score_analysis.R'
+```
+
+Step 5: Next, we commit all changes.
+
+```{bash, eval = FALSE, echo = TRUE}
+# commit command
+git commit -m 'added title to plot illustrating propensity score overlap'
+```
+
+Step 6: Synchronize the changes with the remote repository, so the code is up-to-date on GitHub.
+
+```{bash, eval = FALSE, echo = TRUE}
+# push local changes to remote repository
+git push
+```
+
+{{< pagebreak >}}
+
+## Branching {#sec-branching}
+
+Sometimes it is necessary to implement changes to a document or code without wanting to interfere with a current "`production`" version of a document or code.
+However, a researcher may still want to be able to "`merge`" back potential changes, e.g.
 to integrate new analyses into a script or to develop a new set of features or functions to an application.

### 2023-07-04 manuscript/manuscript.qmd
added key points and made minor modifications here and there

@@
 
-[**Word count:**]{.underline} 3,239 words
+[**Word count:**]{.underline} 3,336 words
 
@@
 
+::: {.callout-note icon="false"}
+## KEY POINTS
+
+-   Transparency and reproducibility are major prerequisites for conducting meaningful real-world evidence (RWE) studies that are fit for decision-making.
+
+-   Regulators, funding agencies and HTA bodies have increasing expectations regarding the provenance, audit trial and sharing of study documents including analytic code as a prerequisite for impactful and credible RWE studies.
+
+-   Version control systems (VCS) for analytic code like Git, GitHub and GitLab have unique advantages for making analytic code and other relevant RWE study documents Findable, Accessible, Interoperable and Reproducible (FAIR).
+
+-   In this manuscript, we discuss and demonstrate how VCS can be used to track changes over the course of a RWE project, collaborate, disseminate and archive analytic code and provide a technical step-by-step tutorial for the practical implementation of reproducible workflows using Git.
+
+-   Given the trend towards open source tools for clinical trial reporting and regulatory submissions, this tutorial addresses a timely topic and aims to encourage new and experienced pharmacoepidemiologists to integrate VCS in their daily work and embrace the advantages of FAIR analytic code sharing.
+:::
+
+{{< pagebreak >}}
+
 # Introduction
@@
 Real-world evidence (RWE) studies that make secondary use of routinely collected health data captured in electronic health records and claims are increasingly being used to inform drug development, regulatory and coverage decisions, as well as clinical practice.[@fdaRWE2018; @eichler2015; @cioms2023; @makady2017] Considering the impact that RWE studies can have on patient healthcare, transparent and reproducible conduct of RWE studies is critically important.
+
 While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], guidance on transparency regarding the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023], SAS or Python.
 Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018; @perez-riverol2016] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community.
-As open source tools like R packages and Shiny dashboards[@pharmaverse2023] become increasingly used for for clinical trial reporting and regulatory submissions[@positPharma], the adoption of version control and code sharing practices is a crucial and timely topic for the field of RWE.
+As open source tools like R packages and Shiny dashboards[@pharmaverse2023] become increasingly used for clinical trial reporting and regulatory submissions[@positPharma], the adoption of version control and code sharing practices is a crucial and timely topic for the field of RWE.
 
@@
 To document the structure and location of files, we can include this, and other critical information about the project, in the `README.md` file.
-This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added aftwerwards by creating a text file with the name `README.md` in the root directory of the repository.
+This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added afterwards by creating a text file with the name `README.md` in the root directory of the repository.
 
@@
 HARPER is a harmonized study protocol template endorsed by the International Society of Pharmacoepidemiology and the International Society of Pharmacoeconomics and Outcomes Research.
-The pre-populated structured study repository template is publicly available and can be `forked`[@GitFork] to initiate a new project (see instructions in the template repository).
+The pre-populated structured study repository template is publicly available and can be `forked`[@GitFork] to initiate a new project (see instructions in the template repository under the provided link).
 
@@
 This step is not needed if the intention of a researcher is just to keep track of local changes and there is no wish to share code, collaborate with others or work across different systems.
-However, most of the times researchers use `Git` for the purpose of collaborating on a project as depicted in @fig-collaborate.
+However, most of the times researchers use `Git` particularly for the purpose of collaborating on a project as depicted in @fig-collaborate.
 
@@
 Through a lockfile, the `renv` package keeps track of the underlying version of R that is being used as well as all R packages, their versions and the installation source (e.g., <https://cran.r-project.org/>).
-To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, the only step to necessary is running the following command in R.
+To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, the only step necessary is running the following command in R.
 
@@
 The main function "`here"` creates paths relative to the top-level of the project directory (in our example the `rwe_study` directory).
-Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any user re-running this script.
+Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any collaborator re-running this script.
 
@@
 library(here)
-icd_codes <- read.csv(here("data", "icd_codes.csv"))
+icd_codes <- read.csv(here("protocol", "icd_codes.csv"))
 ```
@@
 Moreover, when working with remote repositories, it is of utmost importance to ensure that sensitive information (such as passwords or patient-level data) are not accidentally shared.
-To prevent this, we recommend specifying such files in the `.gitignore` file before making the first commit or, more preferably, not storing them in the project directory in the first place.
+To prevent this, we recommend specifying such files in the `.gitignore` file before making the very first commit or, more preferably, not storing them in the project directory in the first place.
 
@@
 Regulators[@fdaRWE2021], funding agencies[@NIH2023;@PCORI2023] and HTA bodies[@kent2021] have increasing expectations regarding the provenance, sharing and audit trial of study documents including analytic code as a prerequisite for impactful and credible RWE studies.
-This manuscript addresses this timely topic and will hopefully motivate and enable new and experienced pharmacoepidemiologists to integrate VCS workflows and code sharing into their daily work.
+Given the trend towards open source tools for clinical trial reporting and regulatory submissions, this tutorial addresses a timely topic and hopefully encourages new and experienced pharmacoepidemiologists to integrate VCS in their daily work and embrace the advantages of FAIR analytic code sharing.
 
@@
 #| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures.
-This commit history illustrates details about the commit (e.g., commit SHA [simple hasing algorithm] and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
+This commit history illustrates details about the commit (e.g., commit SHA [simple hasing algorithm] and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom; line 32 in the code script was added)."
 #| eval: true

### 2023-07-12 manuscript/manuscript.qmd
final compilation for initial submission

@@
 
-Regulators[@fdaRWE2021], funding agencies[@NIH2023;@PCORI2023] and HTA bodies[@kent2021] have increasing expectations regarding the provenance, sharing and audit trial of study documents including analytic code as a prerequisite for impactful and credible RWE studies.
+Regulators[@fdaRWE2021], funding agencies[@NIH2023; @PCORI2023] and HTA bodies[@kent2021] have increasing expectations regarding the provenance, sharing and audit trial of study documents including analytic code as a prerequisite for impactful and credible RWE studies.
 Given the trend towards open source tools for clinical trial reporting and regulatory submissions, this tutorial addresses a timely topic and hopefully encourages new and experienced pharmacoepidemiologists to integrate VCS in their daily work and embrace the advantages of FAIR analytic code sharing.

### 2023-07-12 manuscript/supplement.qmd
added one more amazing Git resource
corrected link

@@
 
-## Example for an end-to-end workflow using Git
+## Example for an end-to-end workflow cycle using Git
 
@@
 
+-   Comprehensive guide on the use of Git in combination with R: <https://happygitwithr.com/>
+
 -   Cheatsheet on most important `Git` commands: <https://education.github.com/git-cheat-sheet-education.pdf>
@@
 
-    -   Division of Pharmacoepidemiology and Pharmacoeconomics: [https://gitlab-scm.partners.org/drugepi](https://gitlab-scm.partners.org/drugepi/harper){.uri}
+    -   Division of Pharmacoepidemiology and Pharmacoeconomics: <https://gitlab-scm.partners.org/drugepi>
 

### 2023-07-12 manuscript/manuscript.qmd
minor correction to Figure 6 size

@@
 #| echo: false
-#| out-width: 105%
+#| out-width: 100%
 

### 2023-11-15 examples/scripts/01_new_user_cohort_query.R
added example code used to create figure 4 and 6

@@ -1,0 +1,2 @@
+# load data example
+data <- smdi_data_complete

### 2023-11-15 examples/scripts/02_descriptives.R

@@ -1,0 +1,1 @@
+# create table 1 example

### 2023-11-15 examples/scripts/03_propensity_score_analysis.R

@@ -1,0 +1,42 @@
+# load data
+library(smdi)
+library(gtsummary)
+library(tidyverse)
+library(here)
+
+# load pre-queried and cleaned cohort data (smdi_data)
+data <- smdi_data_complete
+
+# TEST CHANGES MADE IN LASSO_VARIABLE_SELECTION BRANCH
+
+# step 1: compute propensity of exposure assignment
+exposure_form <- as.formula(paste("exposure ~ ", paste(colnames(data %>% dplyr::select(-exposure)), collapse = " + ")))
+
+exposure_fit <- stats::glm(
+  exposure_form,
+  data = data,
+  family = "binomial"
+)
+
+odds_table <- exposure_fit %>% 
+  gtsummary::tbl_regression(exponentiate = T)
+
+# step 2: compute propensity scores and visualize overlap before matching
+exposure_plot <- data %>% 
+  dplyr::mutate(ps = fitted(exposure_fit))
+
+# plot density/overlap before matching
+ps_plot <- exposure_plot %>% 
+  ggplot2::ggplot(ggplot2::aes(x = ps, fill = factor(exposure))) +
+  ggplot2::geom_density(alpha = .5) +
+  ggplot2::theme_bw() +
+  ggplot2::labs(
+    title = "Propensity score overlap",
+    x = "Pr(exposure)",
+    y = "Density",
+    fill = "Exposed"
+    )
+
+png(filename = here("figures", "Figure_1_propensity_score_overlap.png"), width = 15, height = 10, units = "in", res=600)
+ps_plot + ggplot2::theme(plot.title=element_text(size=70))
+dev.off()

### 2023-11-15 manuscript/manuscript.qmd
implemented reviewer #1 comments/revisions

@@
 [Analytical code sharing statement:]{.underline} This manuscript was written using `Quarto` version 1.3.433 (<https://quarto.org/>) and R version 4.1.2.
-All materials can be found at <https://gitlab-scm.partners.org/drugepi/fair-epi>.
+All materials can be found at <https://gitlab-scm.partners.org/drugepi/fair-epi> or <https://github.com/janickweberpals/fair-epi>.
 Detailed information on packages and versions can be found in the `renv.lock` file in the linked repository.
@@
 
-While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], guidance on transparency regarding the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023], SAS or Python.
+While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], guidance on transparency regarding the actual implementation and analytic steps is still significantly lacking.
 Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018; @perez-riverol2016] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community.
@@
 
+## Creating persistent identifiers
+
+A cornerstone of FAIR principles is the use of persistent identifiers, such as digital object identifiers (DOI), for enabling findability and accessibility of research objects.
+Integrating the concept of persistent identifiers within the context of remote repositories is an important aspect to enhance the traceability and accessibility of code repositories or even specific commits or releases of a study codebase.
+
+To that end, Zenodo[@europeancommission.directorategeneralforresearchandinnovation.2019;@zenodo], a free and open-source platform funded by the European Commission and developed and maintained by European Organization for Nuclear Research (CERN), allows users to share diverse research artifacts.
+The platform seamlessly interfaces with GitHub, enabling the creation of referenceable DOIs for existing code repositories.
+The steps required to obtain a DOI involve the linkage of a Zenodo and a GitHub account[@zenodoGitHub].
+Once linked, a user will be able to select a specific release of a repository which automatically registers a new DOI and a corresponding badge which can be included and referenced in a GitHub README file.
+
 ## Keeping track of R software versions and paths
@@
 Despite the potential initial complexities to get acquainted with the `Git` workflow, distributed VCS and code sharing provide unique advantages, such as a full audit trail of which changes were made by whom, when and why.
-Git makes it easier for users to go back to any snapshot and resolve errors more efficiently.
-Further, remote repositories are ideal platforms to deploy and archive scripts and documents, facilitate collaboration and make it very convenient to share code.
-Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book], Sentinel[@brown2020] and PCORnet[@fleurence2014], which makes shared code interoperable even without direct access to the same data source.
+Remote repositories are ideal platforms to deploy and archive scripts and documents, facilitate collaboration and make it very convenient to share code.
+While the access to the final version of the programming code is already sufficient for external reproducibility, the tracking of changes over time enables a study team to comprehend and resolve errors more efficiently.
+Similar to tracking changes and amendments of a study protocol, an external audience also benefits from a complete audit trail through a better understanding of the documented scientific rationale behind specific changes, the identification of implemented quality control processes (e.g., code reviews) and the tracking of contributions of different collaborators to the codebase.
+
+Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book], Sentinel[@brown2020], PCORnet[@fleurence2014] and ConcePTION [@thurin2021], which makes shared code interoperable even without direct access to the same data source.
+In order to achieve full replicability, authors may alternatively include a sample of dummy data to enable readers to comprehend and run the analytic code.
+Moreover, many larger research data sources have already provided synthetic data using identical formats and variable names while protecting patient privacy (e.g., Medicare Claims Synthetic Public Use Files).\[@gonzal
+
 Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unnecessary redundancies through sharing of code libraries, functions and algorithms.

### 2023-11-22 manuscript/supplement.qmd
added git fetch and merge step
moved reproducible computation environments to supplement as per R#2 comment

@@
 ```{bash, eval = FALSE, echo = TRUE}
-# git pull changes
+# git fetch to see changes without integrating them locally
+git fetch
+
+# git merge to integrate changes to local repository
+get merge
+
+# or directly git pull changes
 git pull
@@
 
+# Reproducible computation environments in R
+
+## Keeping track of R software versions and paths
+
+A major advantage of modern programming languages is that they provide a large pool of open-source packages.
+Packages are compilations of functions that automate frequently used procedures (e.g., querying new user cohorts[@tidyverse] or propensity score matching[@MatchIt]) which expedites the implementation of RWE studies, mitigates copy-paste errors and makes code much more friendly to read and review.
+
+A further step to improve reproducibility when using the R programming language is to ensure that anyone, intending to reproduce analytic results from a `Git` repository, is able to work off the same R package versions and dependencies without having to perform manual installations or path adjustments.
+These requirements are nothing exclusive to R, but apply to any type of programming language.
+
+### Reproducible R computing environments
+
+One way to achieve this is through self-contained computing environments via Docker images which are isolated and standardized environments which include everything needed to run and deploy applications and code regardless of a user's local operating system.[@boettiger2017a; @nüst2020] Another more lightweight, yet effective way is by using the `renv` R package[@renv], which helps creating **re**producible **env**ironments and manages R package dependencies for a given project.
+Through a lockfile, the `renv` package keeps track of the underlying version of R that is being used as well as all R packages, their versions and the installation source (e.g., <https://cran.r-project.org/>).
+To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, the only step necessary is running the following command in R.
+
+```{r}
+renv::restore()
+```
+
+Appending the session info to the end of an R script can provide further context about the dependencies and versions of an analytic script.
+
+```{r}
+sessionInfo()
+```
+
+### Absolute versus relative paths
+
+Another common issue when trying to reproduce analytic code is the use of absolute paths such as in the following example, where we try to call an `icd_code` file from the project's `protocol` sub-directory.
+
+```{r}
+icd_codes <- read.csv('C:/User/Projects/rwe_study/protocol/icd_codes.csv')
+```
+
+The use of absolute paths will make it impossible for any collaborator to run the code without having to manually adapt every single path throughout a repository which is tedious and time consuming.
+A better solution to this is the use of relative paths.
+A popular R package to implement this is the `here` package[@here] which enables file referencing in project-oriented workflows.
+The main function "`here"` creates paths relative to the top-level of the project directory (in our example the `rwe_study` directory).
+Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any collaborator re-running this script.
+
+```{r}
+library(here)
+icd_codes <- read.csv(here("protocol", "icd_codes.csv"))
+```
+
+#
+
 # Supplementary Figures

### 2023-11-22 manuscript/manuscript.qmd
As per R#2 comment: added more commands (fetch, merge, rebase) and moved this as first occurrence in git workflow section, highlighted "time machine" aspect, moved reproducible computation environments to supplement

@@
 
-Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it makes a *snapshot* of a research project repository every time the state of a project is saved.
-For our purposes, a repository can be seen as an isolated project directory which includes all neccessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).
+Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it makes a *snapshot* of a research project repository every time the state of a project is saved and hence can be seen as a "*time machine*" that enables researchers and developers to compare code and repositories across different versions over time.
+For our purposes, a repository can be seen as an isolated project directory which includes all necessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).
 
@@
 
-## Repository structure {#sec-repository-structure}
+In case a users wants to join or contribute to an already existing project, `git clone` works the same way with the only difference that the repository already contains some files and is not empty.
 
-At this step, we now can populate the repository and start working on study-relevant documents and analyses.
+## Repository structure, README file and licenses {#sec-repository-structure}
+
+At this step, we now can populate the repository and start working on study-relevant files and analyses.
 To enhance clarity, it is recommended to use a logical structure and an informative nomenclature for sub-directories and file names (a minimal example is illustrated in @fig-structure).
+
 To document the structure and location of files, we can include this, and other critical information about the project, in the `README.md` file.
 This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added afterwards by creating a text file with the name `README.md` in the root directory of the repository.
+It is critically important to use the `README.md` file to provide essential instructions and information about the repository and the corresponding project.
+This should include a brief summary about the project's background and objectives, instructions on how to install software or manage computation environments and dependencies, details regarding codebooks, important files and the organization of the repository, contribution guidelines, contact information and the project's license to outline the terms under which others can use, modify, and distribute the project.
+Especially for open-source software projects choosing a suitable license is critical for which GitHub provides helpful guidance and information under <https://choosealicense.com/>.
 
@@
 
+### Synchronize changes from a remote repository to a local repository (`pull`, `fetch`, `merge` and `rebase`)  {#pull-fetch-merge}
+
+Especially when working collaboratively with multiple users on a remote repository, it is usually good practice to synchronize the status of one's local repository and the remote repository before making any changes since there could be potential modifications and updates already made by others.
+To avoid any conflicts between two different versions and always work with the most recent version of a file, the first step of the git workflow involves a `git pull` or `git fetch` command.
+
+-   `git fetch` downloads any changes made to a remote repository into a user's local repository.
+However, the command isolates the potentially modified files from the state of the existing local files.
+This is useful since it enables users to safely review changes using `git log` (we will cover this in more detail in @sec-tracking-changes) before integrating those locally.
+To eventually allow the integration of these changes locally, users can do so using the `git merge` or `git rebase` command.
+The difference between these two commands is how the history of the versions between different snapshots of the repository is tracked, for which `git merge` is more common, simpler and less error-prone and `git rebase` is a more advanced feature (for the purpose of this manuscript we will mostly focus on `git merge`).
+
+```{bash}
+# fetch ("download") updated version of a remote repository
+git fetch
+
+# use merge to integrate these changes into local files using merge
+git merge
+```
+
+-   `git pull` combines `git fetch` and `git merge` in one command and directly downloads [and]{.underline} integrates all changes from a remote repository locally.
+
+```{bash}
+# download and integrate content from remote to local
+git pull
+```
+
+Occasionally it can occur that two collaborators made conflicting changes to the same file or even to the same line of code.
+While Git is generally good at automatically integrating (`merging`) new changes, Git will notify the user to resolve a major conflict by editing the conflicting file if Git cannot automatically determine what is correct (e.g., if one collaborator has made changes to a line of code and the other completely deleted it).
+
 ### Stage and commit changes {#sec-stage-and-commit-changes}
 
-As introduced before, Git allows researchers to keep track of changes by making snapshots of the repository every time the state of a project is saved.
-In Git terms, this fundamental step is referred to as a `commit`.
+Git allows researchers to keep track of changes by making snapshots of the repository every time the state of a project is saved.
+In Git terms, this fundamental step is referred to as a `commit` and it enables users to compare code versions across snapshots.
 Generally, it is up to the user when to commit, which files to commit and how the commit should be documented in a `commit message`.
+However, it is best practice to *commit early and often*.
+That is, smaller and more granular commits that reflect a single unit of work (e.g., a change of selected confounders to adjust for) make it easier to comprehend the changes and associated analytic results and revert potential errors.
 
@@
 
-## Tracking changes
+## Tracking changes {#tracking-changes}
 
@@
 
-To that end, Zenodo[@europeancommission.directorategeneralforresearchandinnovation.2019;@zenodo], a free and open-source platform funded by the European Commission and developed and maintained by European Organization for Nuclear Research (CERN), allows users to share diverse research artifacts.
+To that end, Zenodo[@europeancommission.directorategeneralforresearchandinnovation.2019; @zenodo], a free and open-source platform funded by the European Commission and developed and maintained by European Organization for Nuclear Research (CERN), allows users to share diverse research artifacts.
 The platform seamlessly interfaces with GitHub, enabling the creation of referenceable DOIs for existing code repositories.
@@
 Once linked, a user will be able to select a specific release of a repository which automatically registers a new DOI and a corresponding badge which can be included and referenced in a GitHub README file.
-
-## Keeping track of R software versions and paths
-
-A major advantage of modern programming languages is that they provide a large pool of open-source packages.
-Packages are compilations of functions that automate frequently used procedures (e.g., querying new user cohorts[@tidyverse] or propensity score matching[@MatchIt]) which expedites the implementation of RWE studies, mitigates copy-paste errors and makes code much more friendly to read and review.
-
-A further step to improve reproducibility when using the R programming language is to ensure that anyone, intending to reproduce analytic results from a `Git` repository, is able to work off the same R package versions and dependencies without having to perform manual installations or path adjustments.
-These requirements are nothing exclusive to R, but apply to any type of programming language.
-
-### Reproducible R computing environments
-
-One way to achieve this is through self-contained computing environments via Docker images which are isolated and standardized environments which include everything needed to run and deploy applications and code regardless of a user's local operating system.[@boettiger2017a; @nüst2020] Another more lightweight, yet effective way is by using the `renv` R package[@renv], which helps creating **re**producible **env**ironments and manages R package dependencies for a given project.
-Through a lockfile, the `renv` package keeps track of the underlying version of R that is being used as well as all R packages, their versions and the installation source (e.g., <https://cran.r-project.org/>).
-To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, the only step necessary is running the following command in R.
-
-```{r}
-renv::restore()
-```
-
-Appending the session info to the end of an R script can provide further context about the dependencies and versions of an analytic script.
-
-```{r}
-sessionInfo()
-```
-
-### Absolute versus relative paths
-
-Another common issue when trying to reproduce analytic code is the use of absolute paths such as in the following example, where we try to call an `icd_code` file from the project's `protocol` sub-directory.
-
-```{r}
-icd_codes <- read.csv('C:/User/Projects/rwe_study/protocol/icd_codes.csv')
-```
-
-The use of absolute paths will make it impossible for any collaborator to run the code without having to manually adapt every single path throughout a repository which is tedious and time consuming.
-A better solution to this is the use of relative paths.
-A popular R package to implement this is the `here` package[@here] which enables file referencing in project-oriented workflows.
-The main function "`here"` creates paths relative to the top-level of the project directory (in our example the `rwe_study` directory).
-Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any collaborator re-running this script.
-
-```{r}
-library(here)
-icd_codes <- read.csv(here("protocol", "icd_codes.csv"))
-```
 

### 2023-11-22 manuscript/supplement.qmd
smaller adjustment to modified supplement version

@@
 
+Besides the usage of `Git` and remote repositories as a version control systems, there are further relevant aspects that determine the reproducibility of analytic code when using open-source software such a R.
+To complement the introduction and discussion on `Git` in the main manuscript, we here want to elaborate on a few further considerations of computational reproducibility when using R.
+
 ## Keeping track of R software versions and paths
@@
 
-### Reproducible R computing environments
+## Reproducible R computing environments
 
@@
 ```{r}
+#| eval: false
+#| echo: true
+
 renv::restore()
@@
 ```{r}
+#| eval: false
+#| echo: true
+
 sessionInfo()
@@
 
-### Absolute versus relative paths
+## Absolute versus relative paths
 
@@
 ```{r}
+#| eval: false
+#| echo: true
+
 icd_codes <- read.csv('C:/User/Projects/rwe_study/protocol/icd_codes.csv')
@@
 ```{r}
+#| eval: false
+#| echo: true
+
 library(here)
@@
 ```
-
-#
 
@@
 ```
+
+# References {.unnumbered}
+
+::: refs
+:::

### 2023-11-22 manuscript/supplement.qmd
minor wording corrections

@@
 
-Besides the usage of `Git` and remote repositories as a version control systems, there are further relevant aspects that determine the reproducibility of analytic code when using open-source software such a R.
-To complement the introduction and discussion on `Git` in the main manuscript, we here want to elaborate on a few further considerations of computational reproducibility when using R.
+Besides the usage of `Git` and remote repositories as a version control systems, there are further relevant aspects that determine the reproducibility of analytic code when using open-source software such as R.
+To complement the introduction and discussion on `Git` in the main manuscript, we here want to elaborate on a few further considerations addressing the computational reproducibility when using R.
 

### 2023-11-22 manuscript/manuscript.qmd
introduced tagging as per R#2 comment

@@
 
+### Tags (`tag`) {#tags}
+
+A handy `Git` feature to highlight particularly important commits, snapshots, code/software releases or milestones of a project is `tagging`.
+In context of RWE studies, this can be a specific version of an analysis, e.g., the one used for regulatory submission or completion of a manuscript version or revision.
+
+Tags can be annotated with a semantic identifier and tag message, e.g.,
+
+```{bash}
+git tag -a v2.0 -m "Revised (R1) manuscript version"
+```
+
+All created tags can be displayed using the following command:
+
+```{bash}
+git tag
+```
+
 ## Synchronize changes with remote repository
@@
 
-This step is not needed if the intention of a researcher is just to keep track of local changes and there is no wish to share code, collaborate with others or work across different systems.
+Tags, as described in @sec-tags, need to be separately pushed, e.g.,
+
+```{bash}
+git push origin v1.0
+```
+
+In summary, `git push` is not needed if the intention of a researcher is just to keep track of local changes and there is no wish to share code, collaborate with others or work across different systems.
 However, most of the times researchers use `Git` particularly for the purpose of collaborating on a project as depicted in @fig-collaborate.
@@
 To get familiar with more advanced `Git` usage, we compiled a non-exhaustive list of further suggested readings and resources in the **Supplementary Material**.
+
+Besides the usage of `Git` and remote repositories as a version control systems, there are also further relevant aspects that determine the reproducibility of analytic code when using open-source software such as R.
+To complement the introduction and discussion on `Git` in this manuscript, section 2 (*Reproducible computation environments in R*) in the **Supplementary Material** provides a few further considerations addressing the computational reproducibility when using R.
 

### 2023-11-22 manuscript/manuscript.qmd
explicitly point out that we refer to the RStudio IDE

@@
 In this section, we want to give a technical introduction on how Git and remote repositories can be used in RWE studies.
-For the following examples, we focus on workflows utilizing R/RStudio ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as a remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository providers, respectively.
+For the following examples, we focus on workflows utilizing R/the RStudio integrated development environment (IDE) GUI ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as a remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository providers, respectively.
 We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated GUI.

### 2023-11-22 manuscript/manuscript.qmd
added section on security considerations as per R#2 comment

@@
 To do so, users will need to log into their remote repository account and navigate to the section `New repository` (**Supplementary Figure 1**).
-This will take a user to a form to populate basic information of the new project such as the repository name (`rwe_study`) and choose basic settings such as if the repository should be private or public and if `README.md` (see @sec-repository-structure) and `.gitignore` (see @sec-stage-and-commit-changes) files should be automatically created (**Supplementary Figure 2**).
+This will take a user to a form to populate basic information of the new project such as the repository name (`rwe_study`) and choose basic settings such as if the repository should be private or public and if `README.md` (see @sec-repository-structure) and `.gitignore` (see @sec-stage-and-commit-changes and @sec-security) files should be automatically created (**Supplementary Figure 2**).
 
@@
 If any file(s) in the study repository should not be tracked and hence not be synchronized with the remote repository (e.g., because they include confidential information that should not be visible to others like database credentials), these can be specified in the `.gitignore` file which is typically located in the root directory of the repository.
-To create a `.gitignore` file, a user can either auto-generate it upon initialization of the repository on GitHub (**Supplementary Figure 2**) or manually add a text file called `.gitignore` to the root directory.
+To create a `.gitignore` file, a user can either auto-generate it upon initialization of the repository on GitHub (**Supplementary Figure 2**) or manually add a text file called `.gitignore` to the root directory (more information in @sec-security).
 
-### Tags (`tag`) {#tags}
+#### Tags (`tag`) {#sec-tags}
 
@@
 
-## Tracking changes {#tracking-changes}
+## Tracking changes {#sec-tracking-changes}
 
@@
 
+## Security considerations {#sec-security}
+
+When working with Git and remote repositories, several security considerations are crucial to safeguard sensitive information.
+First, it is of utmost importance to ensure that sensitive information (such as personal access tokens, passwords or patient-level data) are not accidentally shared.
+One way to prevent this is to configure a `.gitignore` file, which is a flat text file typically stored in the root directory of the repository.
+Each line in `.gitignore` specifies a file name, sub-directory name or a file name pattern (e.g., "\*.csv" for all files of type csv) that a user intentionally does not want to track and synchronize with a remote repository.
+If it can be avoided, such sensitive information should alternatively not be stored in the project directory in the first place.
+
+```{bash}
+# Example .gitgnore file specyfing which 
+# files, sub-directories and patterns
+# should not be tracked
+
+# password .txt file
+pwd.txt
+
+# Session Data files
+.RData
+.Rhistory
+
+# any type of .csv file
+*.csv
+
+# output directory
+output/
+
+# .Renviron in root directory
+.Renviron
+```
+
+Second, it is good practice to not include passwords or personal access tokens inside an analytic code file but to call these from an external file.
+RStudio, for example, offers the possibility to create environment files (`.Renviron` and `.RProfile`) which are user-controllable text files which can be stored in the project root directory and is automatically called upon an R session start.
+Such a file can be configured to not be tracked using `.gitignore` (as illustrated above) and accessed as show in the following.
+
+```{r}
+# Example: configuration of database access credentials inside an .Renviron file
+db_username = "JoshSmith"
+db_password = "User1234"
+```
+
+These credentials can now be called inside an R script, for example, to access a database (example taken from `genieBPC` R package [@lavery2022]).
+
+```{r}
+# Example how the credentials defined in .Renviron can be used
+# inside an RScript
+set_synapse_credentials(
+  username = Sys.getenv("db_username"), 
+  password = Sys.getenv("db_password")
+  )
+```
+
+Third, it needs to be ensured that access permissions to remote repositories are appropriately configured, granting only necessary access levels to collaborators unless it is intended to share the repository with the public.
+All major remote repositories offer the option to configure a repository to be private or public.
+
 # Discussion
@@
 
-Moreover, when working with remote repositories, it is of utmost importance to ensure that sensitive information (such as passwords or patient-level data) are not accidentally shared.
-To prevent this, we recommend specifying such files in the `.gitignore` file before making the very first commit or, more preferably, not storing them in the project directory in the first place.
-
 Despite the potential initial complexities to get acquainted with the `Git` workflow, distributed VCS and code sharing provide unique advantages, such as a full audit trail of which changes were made by whom, when and why.
@@
 In order to achieve full replicability, authors may alternatively include a sample of dummy data to enable readers to comprehend and run the analytic code.
-Moreover, many larger research data sources have already provided synthetic data using identical formats and variable names while protecting patient privacy (e.g., Medicare Claims Synthetic Public Use Files).\[@gonzal
+Moreover, many larger research data sources have already provided synthetic data using identical formats and variable names while protecting patient privacy (e.g., Medicare Claims Synthetic Public Use Files).[@gonzales2023]
 

### 2023-11-23 manuscript/manuscript.qmd
added details regarding code documentation, codebooks and relevance and challenges of Git in diverse RWE teams (as per R#3 comments)

@@
 
-To document the structure and location of files, we can include this, and other critical information about the project, in the `README.md` file.
+All relevant metadata, instructions and information about a project should be documented in the `README.md` file.
 This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added afterwards by creating a text file with the name `README.md` in the root directory of the repository.
-It is critically important to use the `README.md` file to provide essential instructions and information about the repository and the corresponding project.
-This should include a brief summary about the project's background and objectives, instructions on how to install software or manage computation environments and dependencies, details regarding codebooks, important files and the organization of the repository, contribution guidelines, contact information and the project's license to outline the terms under which others can use, modify, and distribute the project.
+Essential information to include contain a brief summary about the project's background and objectives, instructions on how to install software or manage computation environments and dependencies, important files and the organization of the repository, contribution guidelines, contact information and the project's license to outline the terms under which others can use, modify, and distribute the project.
 Especially for open-source software projects choosing a suitable license is critical for which GitHub provides helpful guidance and information under <https://choosealicense.com/>.
+
+A special aspect about real-world data, such as administrative insurance claims and electronic health records, is that these data are typically not collected for the primary purposes of doing research but rather for reasons of billing or clinical documentation.
+Hence, being transparent about the data provenance, pre-processing steps and operationalization of exposures, outcomes and covariates is of critical importance [@wang2017] and should be part of the repository's metadata.
+This can come in form of codebooks, data dictionaries or other applicable documentation files.
+Applying and documenting naming conventions of files, data structures and variables as developing reproducible and well documented functions or packages as well as inline commenting of programming code can further enhance the transparency and readability of analytic code.[@bové2023improving]
 
@@
 Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unnecessary redundancies through sharing of code libraries, functions and algorithms.
+`Git` was primarily designed for version control and development of analytic code by statisticians and programmers, but may also benefit other members of a RWE generation team through the centralized deployment, tracking and version control of study documents as well as project management tools which have become standard among most established remote repository platforms.
+While `Git` offers substantial benefits for collaboration, reproducibility, and transparency in RWE teams, challenges can arise in its adoption, particularly related to merging conflicts, steep learning curves, and maintaining consistent usage practices across team members.
 

### 2023-11-23 manuscript/manuscript.qmd
minor modifications to figure captions as per R#3 comment

@@
 Hence, being transparent about the data provenance, pre-processing steps and operationalization of exposures, outcomes and covariates is of critical importance [@wang2017] and should be part of the repository's metadata.
-This can come in form of codebooks, data dictionaries or other applicable documentation files.
+This can come in form of codebooks, data dictionaries, literate programming scripts (e.g., Quarto) or other applicable documentation files.
 Applying and documenting naming conventions of files, data structures and variables as developing reproducible and well documented functions or packages as well as inline commenting of programming code can further enhance the transparency and readability of analytic code.[@bové2023improving]
@@
 #| fig-cap: "Minimal example of a transparent repository structure containing relevant study documents.
-Abbreviations: .git = Git sub-directory which includes all files to keep track of changes (created as a result of git init command), renv = R environment sub-directory which contains the project library and other project-specific files and settings needed to manage R package versions and the R environment, renv.lock = R environment lockfile, describing the R version and R package versions used in a project."
+Abbreviations: .git = Git sub-directory which includes all files to keep track of changes (created as a result of git init command), renv = R environment sub-directory which contains the project library and other project-specific files and settings needed to manage R package versions and the R environment, renv.lock = R environment lockfile, describing the R version and R package versions used in a project, *.qmd = Markdown-specific format based on the Quarto open-source scientific and technical publishing system."
 #| eval: true
@@
 #| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures.
-This commit history illustrates details about the commit (e.g., commit SHA [simple hasing algorithm] and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom; line 32 in the code script was added)."
+This commit history illustrates details about the commit (e.g., commit SHA [simple hasing algorithm] and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom; line 32 in the code script was added).
+The unique abbreviated SHA token for this specific change/commit is '9b1a94c' (see 'Commit SHA token' box on the upper right of the figure)."
 #| eval: true

### 2023-11-23 manuscript/manuscript.qmd
final work through manuscript and minor edits

@@
 Git is a free and open source distributed VCS software which was developed in 2005 by the LINUX developer community primarily with the intent to handle large software projects efficiently.[@chacon2014pro] The way to conceptually think about how Git works is that it makes a *snapshot* of a research project repository every time the state of a project is saved and hence can be seen as a "*time machine*" that enables researchers and developers to compare code and repositories across different versions over time.
-For our purposes, a repository can be seen as an isolated project directory which includes all necessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).
+For our purposes, a repository can be thought of as an isolated project directory which includes all necessary files and documents for a given research study ( e.g., protocol, programming code, manuscript, tables, figures, etc.).
 
@@
 In this section, we want to give a technical introduction on how Git and remote repositories can be used in RWE studies.
-For the following examples, we focus on workflows utilizing R/the RStudio integrated development environment (IDE) GUI ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as a remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository providers, respectively.
+For the following examples, we focus on workflows utilizing the RStudio integrated development environment (IDE) GUI ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/https://posit.co/download/rstudio-desktop/)) and GitHub (<https://github.com/>) as a remote repository, although the general concepts are fairly similar and easily transferable to other programming languages and remote repository platforms, respectively.
 We will show examples for both command line interface (CLI) prompts as well as for RStudio's integrated GUI.
@@
 This can come in form of codebooks, data dictionaries, literate programming scripts (e.g., Quarto) or other applicable documentation files.
-Applying and documenting naming conventions of files, data structures and variables as developing reproducible and well documented functions or packages as well as inline commenting of programming code can further enhance the transparency and readability of analytic code.[@bové2023improving]
+Developing reproducible and well documented functions or packages as well as inline commenting of programming code can further enhance the transparency and readability of analytic code.[@bové2023improving]
 
@@
 Every time a user makes a commit to create a snapshot of the work, `Git` creates a unique hash (a 40-character string created by a **simple hashing algorithm** \[SHA\]) which can be seen as an "ID" of the snapshot which enables users to comprehend every change made and revert back to any snapshot of the repository.
-@fig-stage-commit-RStudio-gui illustrates the equivalent steps in RStudio GUI.
+@fig-stage-commit-RStudio-gui illustrates the equivalent steps in the RStudio GUI.
 
@@
 
-#### Tags (`tag`) {#sec-tags}
+### Tags (`tag`) {#sec-tags}
 
 A handy `Git` feature to highlight particularly important commits, snapshots, code/software releases or milestones of a project is `tagging`.
-In context of RWE studies, this can be a specific version of an analysis, e.g., the one used for regulatory submission or completion of a manuscript version or revision.
+In context of RWE studies, this can be a specific version of an analysis, e.g., the one used for a regulatory submission or the completion of a manuscript version or revision.
 
@@
 ```{bash}
-git push origin v1.0
+git push origin v2.0
 ```
@@
 However, most of the times researchers use `Git` particularly for the purpose of collaborating on a project as depicted in @fig-collaborate.
-
-### Synchronize remote to local (`pull`)
-
-In case collaborators have also made modification to their local files and pushed them to the remote repository, these changes can be easily synchronized with our local copy via the `pull` command.
-
-```{bash}
-# pull remote repository version of code to local
-git pull
-```
-
-Occasionally it can occur that two collaborators made conflicting changes to the same file or even to the same line of code.
-While Git is generally good at automatically integrating (`merging`) new changes, Git will notify the user to resolve a major conflict by editing the conflicting file if Git cannot automatically determine what is correct (e.g., if one collaborator has made changes to a line of code and the other completely deleted it).
 
@@
 # files, sub-directories and patterns
-# should not be tracked
+# that should not be tracked
 
@@
 
-# .Renviron in root directory
+# .Renviron file
 .Renviron
@@
 # Example how the credentials defined in .Renviron can be used
-# inside an RScript
+# inside an R script
 set_synapse_credentials(
@@
 
-Third, it needs to be ensured that access permissions to remote repositories are appropriately configured, granting only necessary access levels to collaborators unless it is intended to share the repository with the public.
+Third, it needs to be ensured that access permissions to remote repository platforms are appropriately configured, granting only necessary access levels to collaborators unless it is intended to share the repository with the public.
 All major remote repositories offer the option to configure a repository to be private or public.
@@
 
-Besides the usage of `Git` and remote repositories as a version control systems, there are also further relevant aspects that determine the reproducibility of analytic code when using open-source software such as R.
+Besides the usage of `Git` and remote repositories as version control systems, there are also further relevant aspects that determine the reproducibility of analytic code when using open-source software such as R.
 To complement the introduction and discussion on `Git` in this manuscript, section 2 (*Reproducible computation environments in R*) in the **Supplementary Material** provides a few further considerations addressing the computational reproducibility when using R.
@@
 Finally, code sharing can enable researchers to learn from each other, build trust[@orsini2020], expedite the uptake of novel methodologies and knowledge exchange and decrease unnecessary redundancies through sharing of code libraries, functions and algorithms.
-`Git` was primarily designed for version control and development of analytic code by statisticians and programmers, but may also benefit other members of a RWE generation team through the centralized deployment, tracking and version control of study documents as well as project management tools which have become standard among most established remote repository platforms.
+`Git` was primarily designed for version control and development of analytic code by statisticians and programmers, but may also benefit other members of a RWE generation team through the centralized deployment, tracking and version control of study documents as well as project management tools which have become standard among established remote repository platforms.
 While `Git` offers substantial benefits for collaboration, reproducibility, and transparency in RWE teams, challenges can arise in its adoption, particularly related to merging conflicts, steep learning curves, and maintaining consistent usage practices across team members.
@@
 #| label: fig-RStudio-gui
-#| fig-cap: "Steps to clone a remote repository using the RStudio graphical user interface."
+#| fig-cap: "Steps to clone a remote repository using the RStudio integrated development environment (IDE) graphical user interface (GUI)."
 #| eval: true
@@
 #| label: fig-stage-commit-RStudio-gui
-#| fig-cap: "RStudio provides a graphical user interface as an alternative to command line prompts to interact with Git and to perform command equivalent to git add, git commit and git push/pull (red circles).
-This view can be accessed in the right upper pane of RStudio under Git > Diff.
+#| fig-cap: "The RStudio integrated development environment (IDE) provides a graphical user interface (GUI) as an alternative to command line prompts to interact with Git and to perform command equivalent to git add, git commit and git push/pull (red circles).
+This view can be accessed in the right upper pane of the RStudio IDE under Git > Diff.
 In this example, an R script with the title '03_propensity_score_analysis.R' is selected and staged.
 All changes performed in this file can be viewed in the window below with green highlighted lines indicating additions and red highlighted lines deletions from the prior version.
-In this case, a title was added to a propensity score plot which is also expressed in the commit message."
+In this example, a title was added to a propensity score plot which is also expressed in the commit message."
 #| eval: true

### 2023-11-23 manuscript/manuscript.qmd
adjusted word count
corrected typo

@@
 
-[**Word count:**]{.underline} 3,336 words
+[**Word count:**]{.underline} 4,293 words
 
@@
 
-In case a users wants to join or contribute to an already existing project, `git clone` works the same way with the only difference that the repository already contains some files and is not empty.
+In case a user wants to join or contribute to an already existing project, `git clone` works the same way with the only difference that the repository already contains some files and is not empty.
 

### 2023-11-29 manuscript/manuscript.qmd
implemented SVW comments to manuscript and fine-tuned responses

@@
 
-[**Word count:**]{.underline} 4,293 words
+[**Word count:**]{.underline} 4,309 words
 
@@
 Remote repositories are ideal platforms to deploy and archive scripts and documents, facilitate collaboration and make it very convenient to share code.
-While the access to the final version of the programming code is already sufficient for external reproducibility, the tracking of changes over time enables a study team to comprehend and resolve errors more efficiently.
+While the access to the final version of the programming code along with the source data is sufficient for external reproducibility, the tracking of changes over time enables a study team to comprehend and resolve errors more efficiently.
 Similar to tracking changes and amendments of a study protocol, an external audience also benefits from a complete audit trail through a better understanding of the documented scientific rationale behind specific changes, the identification of implemented quality control processes (e.g., code reviews) and the tracking of contributions of different collaborators to the codebase.
 
-Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book], Sentinel[@brown2020], PCORnet[@fleurence2014] and ConcePTION [@thurin2021], which makes shared code interoperable even without direct access to the same data source.
-In order to achieve full replicability, authors may alternatively include a sample of dummy data to enable readers to comprehend and run the analytic code.
-Moreover, many larger research data sources have already provided synthetic data using identical formats and variable names while protecting patient privacy (e.g., Medicare Claims Synthetic Public Use Files).[@gonzales2023]
+Although data sharing is unfortunately often prohibited due to privacy and legal reasons, many databases have been mapped to common data models like OMOP[@ohdsi2019book], Sentinel[@brown2020], PCORnet[@fleurence2014] and ConcePTION[@thurin2021], which makes shared code interoperable even without direct access to the same data source.
+Moreover, many larger research data sources have already provided synthetic data using identical formats and variable names while protecting patient privacy (e.g., Medicare Claims Synthetic Public Use Files).[@gonzales2023] Authors may also include a sample of dummy data to give readers a better understanding of the underlying source code and enhance transparency, but, by definition, reproducibility can only be achieved by having access to real data.
 
